### 2016-02-25 paper.tex
Scaffolding for SPARK16.

\def\year{2016}\relax
%File: formatting-instruction.tex
\documentclass[letterpaper]{article}
\usepackage{aaai16}
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}

\usepackage[pdftex]{hyperref}
\usepackage{amsmath}
\usepackage{color}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage{ccicons}
\usepackage{todonotes}
\usepackage{amsfonts}

\frenchspacing
\setlength{\pdfpagewidth}{8.5in}
\setlength{\pdfpageheight}{11in}
\pdfinfo{
/Title (Insert Your Title Here)
/Author (Put All Your Authors Here, Separated by Commas)}
\setcounter{secnumdepth}{0}

\newcommand{\aux}{\mathit{aux}}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\cA}{\mathcal{A}}
\newcommand{\tit}[1]{\textit{#1}}
\newcommand{\IFF}{\textit{iff}}
\newcommand{\IF}{\textit{if}}
\newcommand{\PCF}{\mathit{PCF}}
\newcommand{\walk}{\mathit{walk}}
\newcommand{\bike}{\mathit{bike}}
\newcommand{\car}{\mathit{car}}
\newcommand{\public}{\mathit{public}}
\newcommand{\taxi}{\mathit{taxi}}
\newcommand{\gas}{\mathit{gas}}
\newcommand{\wait}{\mathit{wait}}
\newcommand{\SUM}{\mathit{sum}}
\newcommand{\MAX}{\mathit{max}}
\newcommand{\MIN}{\mathit{min}}
\newcommand{\AVG}{\mathit{avg}}

\newcommand{\secref}[1]{Section~\ref{sec:#1}}
\newcommand{\figref}[1]{Figure~\ref{fig:#1}}
\newcommand{\tblref}[1]{Table~\ref{tbl:#1}}
\newcommand{\algref}[1]{Algorithm~\ref{alg:#1}}
\newcommand{\thmref}[1]{Theorem~\ref{thm:#1}}
\newcommand{\propref}[1]{Proposition~\ref{prop:#1}}
\newcommand{\corref}[1]{Corollary~\ref{cor:#1}}

 \begin{document}
% The file aaai.sty is the style file for AAAI Press 
% proceedings, working notes, and technical reports.
%
\title{On Extensibility and Personalizability of Multi-Modal Trip Planning}
\author{AAAI Press\\
Association for the Advancement of Artificial Intelligence\\
2275 East Bayshore Road, Suite 160\\
Palo Alto, California 94303\\
}
\maketitle
\begin{abstract}
	In the setting of multi-modal trip planning, it is essential for
	the planner to allow users to upload auxiliary data of interest,
	let them express personal constraints and preferences,
	and compute optimal trips accordingly.
	In this work, we designed and implemented a multi-modal trip
	planning framework that is both extensible and personalizable.
	%We conducted a user study on the feasibility of the features of
	%our planner, and showed that the features are desirable and have
	%great potential value in mobility marketplace. 
	To this end, we developed a hipergraph-based trip planner that 
	accommodates constraints specified as a theory in the 
	\tit{linear temporal logic} and preferences as a 
	user-specific \tit{preferential cost function} 
	to compute optimal routes using the A* algorithm.
\end{abstract}

\section{Introduction}
%% \begin{itemize}
%% 	\setlength\itemsep{1pt}
%% 	\item What's the problem: to design and implement a trip planning
%% 				system that is multi-modal, extensible and personalizable.
%% 	\item Why is it important: a story-telling example here should suffice.
%% 	\item Why is it hard: not straightforward how to represent and reason
%% 				about user constraints and preferences, etc.
%% 	\item How did we solve it: allowing the user to upload data and specify
%% 				constraints and preferences as LTL formulas and preferential
%% 				cost functions, respectively.  Then, these constraints and
%% 				preferences are integrated into the A* search.
%% 	\item How did we evaluate the features and what are the results:
%% 				ceteris paribus experimentation setup on AMT and results
%% \end{itemize}

\begin{itemize}
	\item What's the problem: personalized trip planning: find trips that
	  are optimal in the eyes of a user, who doesn't care about mode or
	  provider boundaries, and whose preferences may be more faceted than
	  ``fastest'' or ``cheapest''.
	\item Why is it hard: the computational complexity of searching
	  combinations of multiple modes that can be combined freely is high
	  (PSPACE-complete\cite{}) which is why existing trip planner exploit problem
	  structure (such as only one mode per trip) and predefined
	  preferences (optimization metrics)
	\item How did we solve it: Using a high-performance graph search
	  engine\cite{}, we implemented a multi-modal trip planner that uses pure
	  graph-search. This allows us to flexibly combine various modes
	  (e.g., car, train, cab, walk) and declarative change soft and hard
	  constraints. The planner can take constraints expressed in linear
	  temporal logic to constrain the search space (e.g., never bike after
	  transit, i.e., I don't want to take my bike on the bus or train).
	  The planner uses a weighted sum over several metrics (e.g., time
	  overall, time spend on transit, cost) which can be re-weighted based
	  on different user preferences. Finally the planner allows the user
	  to upload {\em new} mapping data over which to express hard and/or
	  soft constraints. For instance, a user might upload a map of crime
	  in the city, and ask the trip planner to never walk at night through
	  an area where crime is frequent.
%	\item How did we evaluate the features and what are the results:
%	  ceteris paribus experimentation setup on Amazon Mechanical 
%		Turk (MTurk\footnote{\url{http://mturk.amazon.com}}) and results, 
%		asking users to pick between two different trips, one being limited by the
%	  capabilities of existing planners, and one exhibiting features only
%	  available in our trip planner.
\end{itemize}

\section{Related Work}
\begin{itemize}
	\setlength\itemsep{1pt}
	\item Summarize related work on multi-modal trip planning.
	\item Summarize related work on constraints and preferences in planning.
		\begin{itemize}
			\setlength\itemsep{0pt}
			\item LTL in planning.
			\item Numeric/ordinal/qualitative preferences in planning.
		\end{itemize}
	\item Summarize related work on VOT (Value-of-Time).
\end{itemize}
Constraints and preferences in planning have been studied in the 
literature.
Son and Pontelli presented a declarative preference language, 
$\mathcal{PP}$, to represent ordinal preferences (e.g., basic desires,
atomic preferences and general preferences) between trajectories of 
a planning problem~\cite{son2004planning}.
Bienvenu et al. proposed a first-order preference language, $\mathcal{LPP}$, 
extending $\mathcal{PP}$ by allowing the specification of quantified qualitative
preferences~\cite{bienvenu2011specifying}.
Bast et al. surveyed algorithms, including Dijkstra and A*, in the setting of
multi-modal transportation networks ~\cite{bast2015route}.
Modesti and Sciomachen introduced an ad hoc utility function for weighing
the arcs both with their time and monetary cost~\cite{modesti1998utility}.
Economic models capturing the trade-off between travel time and travel
cost have been presented~\cite{antoniou2007methodology}.

\section{Extensibility}
\begin{itemize}
	\setlength\itemsep{1pt}
	\item Explain in detail what extensibility means.
	\item Describe the input.
	\item Describe how we incorporate the auxiliary data into the planner.
\end{itemize}
Allowing users to upload their own data sets of interests is an important
step towards customization of a trip planner.
We designed a framework where a user can load the data sets (e.g., crime
statistics and pollution data) into the planner,
and the planner will compute an optimal route accordingly.

\subsection{Auxiliary Data}
The user-created data are the auxiliary data that is represented as pairs
of latitude and longitude degrees.
To merge these lat-long pairs into the planner, we performed a neighborhood
search to calculate the total score of auxiliary data for each 
lat-long pair already in our planner.
It might be of strong interest to some user for our planning
system to take care of criminal statistics so that some level of safety
of the resulting routes is guaranteed.
For instance, a user traveling through the downtown area of San Francisco
around midnight may want to upload a data set of crimes (cf. \figref{sf_crime}), 
and express her constraints and preferences in hope of a preferred trip plan.

\begin{figure}[!ht]
  \centering
    \includegraphics[width=0.4\textwidth]{figs/sub_sf_crime_July2015.png}
  \caption{Crime rates in San Francisco\label{fig:sf_crime}}
\end{figure}

Formally, let $A$ be the auxiliary set of lat-long pairs, 
$N$ a node in our planner and $r$ the effective radius.
We first compute the set $P$ of auxiliary points such that
$P \subseteq A$ and $\dist(p,N)<r$ for all $p \in P$.
Then, the total score of auxiliary data for $N$ with respect to $P$ and $r$,
denoted by $S_\aux(N,P,r)$, is computed as follows.

\begin{equation}
	S_\aux(N,P,r) = \sum_{\substack{p \in P}} \Big(1 - \frac{\dist(p,N)}{r}\Big)
\end{equation}

\section{Personalizability}
Personalizability consists of two aspects: constraints and preferences.
From the viewpoint of the planner,
constraints, also referred to as hard constraints, are statements that the planner
has to satisfy during the planning process; whereas preferences, also called
soft constraints, are specifications that the planner will need to optimize.
We formulated constraints using linear temporal logic (LTL) and preferences as
a preferential cost function (PCF), and implemented our planner leveraging the
widely-used graph search algorithm the A*.

\subsection{Constraints}
\begin{itemize}
	\setlength\itemsep{1pt}
	\item Describe syntax and semantics of LTL in the setting of trip planning.
	\begin{itemize}
		\setlength\itemsep{0pt}
		\item Specifically, we need to point out the ``after" temporal connective,
					and we need to describe the semantics: what is a plan, why we can
					eliminate actions and put them as state attributes instead, etc.
	\end{itemize}
\end{itemize}
As constraints in the setting of trip planning are often declarative and
temporal, our choice of LTL is straightforward.
We now give a brief review of linear temporal logic (LTL).
Let $f$ be a propositional formula over a finite set $L$ of Boolean variables.  
LTL formulas are defined recursively as follows.
\begin{equation}
	\varphi = f | \varphi_1 \land \varphi_2 | \varphi_1 \lor \varphi_2 | \neg \varphi | 
		\bigcirc \varphi |	\Box \varphi | \Diamond \varphi | \varphi_1 \cA \varphi_2
\end{equation}
Note that we have $\varphi_1 \cA \varphi_2$, and it means that
``$\varphi_2$ holds right after $\varphi_1$ holds."

An intuitive constraint in trip planning could be ``I will never drive after biking or
taking the public transit."
In LTL, such constraint can be translated into an LTL formula
\begin{align*}
	\psi = ((M=b) \lor (M=p)) \,\cA\, (\Box (\neg (M=c))).
\end{align*}

As the actions in trip planning is limited to taking different transportation modes,
in our definition of the semantics of LTL
these actions are subsumed into the interpretations of $L$, or \tit{states}.
The semantics of LTL is defined with regard to trajectories of states. 
Let $\sigma$ be a trajectory of states $S_0,a_1,S_1,\ldots,a_n,S_n$, and
$\sigma[i]$ a suffix $S_i, a_{i+1}, S_{i+1}, \ldots,a_n,S_n$.  We have
\begin{align*}
	\sigma \models f \;\; &\IFF \;\; S_0 \models f,\\
	\sigma \models \varphi_1 \land \varphi_2 \;\; &\IFF \;\; \sigma \models \varphi_1 \; and \; \sigma \models \varphi_2,\\
	\sigma \models \varphi_1 \lor \varphi_2 \;\; &\IFF \;\; \sigma \models \varphi_1 \; or \; \sigma \models \varphi_2,\\
	\sigma \models \neg \varphi \;\; &\IFF \;\; \sigma \not \models \varphi,\\
	\sigma \models \bigcirc \varphi \;\; &\IFF \;\; \sigma[1] \models \varphi,\\
	\sigma \models \Box \varphi \;\; &\IFF \;\; \forall 0 \leq i \leq n (\sigma[i] \models \varphi),\\
	\sigma \models \Diamond \varphi \;\; &\IFF \;\; \exists 0 \leq i \leq n (\sigma[i] \models \varphi),\\
	\sigma \models \varphi_1 \cA \varphi_2 \;\; &\IFF \;\; \forall 0 \leq i < n (\IF \; \sigma[i] \models \varphi_1, \sigma[i+1] \models \varphi_2).
\end{align*}

For example, we are given an LTL constraint $\psi$ and three trajectories ($\sigma_1$, $\sigma_2$ and $\sigma_3$)
as shown in \figref{trjs}.
Clearly, we have $\sigma_1 \models \psi$ because, after public transit in $S_2$ and $S_3$, 
traveling by car has never been taken place.
Moreover, we have $\sigma_2 \not \models \psi$ because we have $M=c$ hold in $S_6$ and $S_7$
after having $M=p$ hold in $S_2$ and $S_5$.
Finally, we know $\sigma_3 \models \psi$, as the mode is always neither biking nor public transit.

\begin{figure}[!ht]
  \centering
    \includegraphics[width=0.5\textwidth]{figs/trajectories.pdf}
  \caption{State transition diagram\label{fig:trjs}}
\end{figure}

\subsection{Preferences}
\begin{itemize}
	\setlength\itemsep{1pt}
	\item Describe our preferential cost function (PCF).
	\begin{itemize}
		\setlength\itemsep{0pt}
		\item Describe why and how we break down time and money.
		\item Describe why and how we design coefficients to these time and money pieces.
		\item Describe why and how we design the two ratios: dollars per hour and dollars per aux.
	\end{itemize}
\end{itemize}

A state is described as a set of \tit{state variables}.
The state variables of a state $S$ include the transportation mode $M$ that led to $S$,
time spent $T$ so far per mode (e.g., $T_\bike$ for biking and $T_\public$ for
public transit), fare $D$ spent so far per mode (e.g., $D_\gas$ for driving and
$D_\taxi$ for taking a cab), and variables related to the auxiliary data once uploaded.
These extra data related variables are metrics such as the sum ($A_\SUM$),
the maximum ($A_\MAX)$, the minimum ($A_\MIN$), and the average ($A_\AVG$) data along the path.

We focused on weighted functions over state variables and
designed the cost function, called preferential cost function (PCF), that guides the
graph-based search engine in our trip planner as follows.
\begin{equation}
	\begin{aligned}
		\PCF(S) =& \beta_1 * (\alpha_1 \cdot T_\walk + \alpha_2 \cdot T_\wait + \ldots) \\
								&+ (D_\gas + D_\public + \ldots) \\
								&+ \beta_2 * (A_\SUM + \ldots),
	\end{aligned}
\end{equation}
where $\alpha_i$ are real numbers representing the relations among different time pieces,
and $\beta_1$ ($\beta_2$) is the ratio that essentially describes how much in dollars a user would pay to
save an hour (an auxiliary data, respectively).

\subsubsection{Preference Elicitation}
To gather these coefficients ($\alpha_i$ and $\beta_i$) in our $\PCF$, we designed interface to
elicit these numbers from the user.

\subsection{Reasoning with Constraints and Preferences}
\begin{itemize}
	\setlength\itemsep{1pt}
	\item Describe why and how we integrate constraints and preferences into A*.
\end{itemize}

We leveraged the widely-used A* search algorithm on top of our high-performance graph search
engine.  The A* algorithm incorporates the following cost function.
\begin{equation}
	f(S) = g(S) + h(S),
\end{equation}
where $g(S)$ is the cost of an optimal trip from the initial state to $S$, and
$h(S)$ is an admissible estimate of the cost of an optimal trip from $S$ to goal.

To prune the search space, we check satisfiability of the temporal constraints in LTL
at expansion of the search tree.
To guide the search engine, we set $g(S)=\PCF(S)$ and $h(S)$ the minimum estimate among
all available modes in $S$.

%\section{User Study}
%\begin{itemize}
%	\setlength\itemsep{1pt}
%	\item Describe the set up of our experiments on AMT and explain why: because we want to evaluate the
%				feasibility of the three features.
%	\item Describe the results.
%\end{itemize}
\section{Implementation}

\begin{itemize}
	\setlength\itemsep{1pt}
	\item Describe the system structure.
	\item Describe the input to the system.
	\item Describe the output from the system.
\end{itemize}

\section{Conclusion and Future Work}

\bibliographystyle{aaai}
\bibliography{refs}

\end{document}

### 2016-03-01 paper.tex
Updated Abstract and Introduction.

@@
 \pdfinfo{
-/Title (Insert Your Title Here)
+/Title (On Extensibility and Personalizability of Multi-Modal Trip Planning)
 /Author (Put All Your Authors Here, Separated by Commas)}
@@
 \title{On Extensibility and Personalizability of Multi-Modal Trip Planning}
-\author{AAAI Press\\
-Association for the Advancement of Artificial Intelligence\\
-2275 East Bayshore Road, Suite 160\\
-Palo Alto, California 94303\\
+\author{
+	Xudong Liu\\
+	Department of Computer Science\\
+	University of Kentucky\\
+	329 Rose Street\\
+	Lexington, KY 40506\\
+	USA
+	\And
+	PARC People\\
+	Palo Alto Research Center\\
+	3333 Coyote Hill Road\\
+	Palo Alto, CA 94304\\
+	USA
 }
@@
 \begin{abstract}
-	In the setting of multi-modal trip planning, it is essential for
-	the planner to allow users to upload auxiliary data of interest,
-	let them express personal constraints and preferences,
-	and compute optimal trips accordingly.
-	In this work, we designed and implemented a multi-modal trip
-	planning framework that is both extensible and personalizable.
-	%We conducted a user study on the feasibility of the features of
-	%our planner, and showed that the features are desirable and have
-	%great potential value in mobility marketplace. 
-	To this end, we developed a hipergraph-based trip planner that 
-	accommodates constraints specified as a theory in the 
-	\tit{linear temporal logic} and preferences as a 
-	user-specific \tit{preferential cost function} 
-	to compute optimal routes using the A* algorithm.
+	Trip planning is a practical task that has drawn
+	extensive attention from the AI planning and scheduling community
+	and the industrial/commercial sectors.
+	In this paper, we consider the setting of the \tit{multi-modal}
+	trip planning, where users can exploit different transportation
+	modes, such as walking, biking, public transit, and taxi.
+	In such a context, it would be of benefit if the
+	user was able to extend the \tit{cost base}, including traveling time 
+	and fare, of the planner, and to personalize the planner according
+	to her own constraints and preferences.
+	To this end, we designed and developed a hipergraph-based multi-modal
+	trip planner that allows users to upload auxiliary cost metrics (e.g.,
+	crime rates), to specify constraints as a theory in the 
+	\tit{linear temporal logic}, and to express preferences as a 
+	\tit{preferential cost function}.
 \end{abstract}
@@
 \section{Introduction}
-%% \begin{itemize}
-%% 	\setlength\itemsep{1pt}
-%% 	\item What's the problem: to design and implement a trip planning
-%% 				system that is multi-modal, extensible and personalizable.
-%% 	\item Why is it important: a story-telling example here should suffice.
-%% 	\item Why is it hard: not straightforward how to represent and reason
-%% 				about user constraints and preferences, etc.
-%% 	\item How did we solve it: allowing the user to upload data and specify
-%% 				constraints and preferences as LTL formulas and preferential
-%% 				cost functions, respectively.  Then, these constraints and
-%% 				preferences are integrated into the A* search.
-%% 	\item How did we evaluate the features and what are the results:
-%% 				ceteris paribus experimentation setup on AMT and results
-%% \end{itemize}
+Trip planning, as an application of planning and scheduling,
+has seen substansive implementations by researchers and
+developers\cite{bast2015route}.
+Some of the planning systems are \tit{multi-modal}; that is,
+combining distinct transportation modes, the trip planners
+compute optimal routes from sources to destinations.
+This notion of ``optimality" generally refers to the computed routes
+having minimal total time or total fare.
 
-\begin{itemize}
-	\item What's the problem: personalized trip planning: find trips that
-	  are optimal in the eyes of a user, who doesn't care about mode or
-	  provider boundaries, and whose preferences may be more faceted than
-	  ``fastest'' or ``cheapest''.
-	\item Why is it hard: the computational complexity of searching
-	  combinations of multiple modes that can be combined freely is high
-	  (PSPACE-complete\cite{}) which is why existing trip planner exploit problem
-	  structure (such as only one mode per trip) and predefined
-	  preferences (optimization metrics)
-	\item How did we solve it: Using a high-performance graph search
-	  engine\cite{}, we implemented a multi-modal trip planner that uses pure
-	  graph-search. This allows us to flexibly combine various modes
-	  (e.g., car, train, cab, walk) and declarative change soft and hard
-	  constraints. The planner can take constraints expressed in linear
-	  temporal logic to constrain the search space (e.g., never bike after
-	  transit, i.e., I don't want to take my bike on the bus or train).
-	  The planner uses a weighted sum over several metrics (e.g., time
-	  overall, time spend on transit, cost) which can be re-weighted based
-	  on different user preferences. Finally the planner allows the user
-	  to upload {\em new} mapping data over which to express hard and/or
-	  soft constraints. For instance, a user might upload a map of crime
-	  in the city, and ask the trip planner to never walk at night through
-	  an area where crime is frequent.
-%	\item How did we evaluate the features and what are the results:
-%	  ceteris paribus experimentation setup on Amazon Mechanical 
-%		Turk (MTurk\footnote{\url{http://mturk.amazon.com}}) and results, 
-%		asking users to pick between two different trips, one being limited by the
-%	  capabilities of existing planners, and one exhibiting features only
-%	  available in our trip planner.
-\end{itemize}
+However, in the eyes of a user, it may be more faceted than just
+``fastest" or ``cheapest."
+For instance, for a college student who specifies that she will
+only walk or take public transit in a trip from Palo Alto to
+San Francisco, the computed plan is not necessarily the fastest
+(e.g., taking a cab could be faster)
+or the cheapest (e.g., walking all the way has no fare).
+This happens when a user tells the planner her hard constraints,
+called \tit{constraints}.  
+The planner then needs to either satisfy the 
+constraints in the search process or return failure because they are
+over-restrictive.
+Moreover, the user might want to further customize the planner by
+describing soft constraints, called \tit{preferences}.
+For example, an agent wants to travel from school to
+downtown, and prefers biking to taking the bus.
+Thus, a trip with more biking than bus may be considered better for
+the agent than the one with more bus than biking.
+In this case, the planner will need to accommodate user preferences
+whenever possilbe in the search of optimal solutions.
 
-\section{Related Work}
-\begin{itemize}
-	\setlength\itemsep{1pt}
-	\item Summarize related work on multi-modal trip planning.
-	\item Summarize related work on constraints and preferences in planning.
-		\begin{itemize}
-			\setlength\itemsep{0pt}
-			\item LTL in planning.
-			\item Numeric/ordinal/qualitative preferences in planning.
-		\end{itemize}
-	\item Summarize related work on VOT (Value-of-Time).
-\end{itemize}
-Constraints and preferences in planning have been studied in the 
-literature.
+Representing and reasoning about constraints and preferences are
+fundamental to decision making in automated planning and scheduling
+in artificial intelligence.
 Son and Pontelli presented a declarative preference language, 
@@
 atomic preferences and general preferences) between trajectories of 
-a planning problem~\cite{son2004planning}.
+a planning problem\cite{son2004planning}.
 Bienvenu et al. proposed a first-order preference language, $\mathcal{LPP}$, 
 extending $\mathcal{PP}$ by allowing the specification of quantified qualitative
-preferences~\cite{bienvenu2011specifying}.
-Bast et al. surveyed algorithms, including Dijkstra and A*, in the setting of
-multi-modal transportation networks ~\cite{bast2015route}.
+preferences\cite{bienvenu2011specifying}.
 Modesti and Sciomachen introduced an ad hoc utility function for weighing
-the arcs both with their time and monetary cost~\cite{modesti1998utility}.
-Economic models capturing the trade-off between travel time and travel
-cost have been presented~\cite{antoniou2007methodology}.
+the arcs both with their time and fare\cite{modesti1998utility}.
+This relationship between time and fare in the setting of transportation
+was studied, and economic models capturing the trade-off between the two
+was presented\cite{antoniou2007methodology}.
+
+However, relatively limited effort has been devoted to designing and
+implementing real-world multi-modal trip planners that captures user
+constraints and preferences over the \tit{cost base}, possibly
+extended from the user with auxiliary \tit{cost metrics},
+such as crime rates and pollution statistics.
+
+Using a high-performance graph search
+engine\cite{zhou2011dynamic}, we designed and implemented a 
+multi-modal trip planner that uses pure
+graph-search. This allows us to flexibly combine various modes
+(e.g., walk, bike, car, public transit, and taxi) and 
+declaratively change constraints and preferences. 
+The planner allows the user
+to upload \tit{new} mapping data over which to express
+constraints and preferences.  For instance, a user might upload a map of crime
+in the city, and ask the trip planner to avoid areas where crime is frequent.
+Then, the planner takes constraints expressed in \tit{linear
+temporal logic} to constrain the search space (e.g., never bike after
+transit, and never walk through bad neighborhoods).
+Finally, the planner uses a weighted sum, called \tit{preferential cost function}, 
+over several cost metrics (e.g., time spend biking, fare on public transit, 
+and overall crimes walking through) which can be re-weighted based
+on different user preferences.
+
+%\begin{itemize}
+%	\item What's the problem: personalized trip planning: find trips that
+%	  are optimal in the eyes of a user, who doesn't care about mode or
+%	  provider boundaries, and whose preferences may be more faceted than
+%	  ``fastest'' or ``cheapest''.
+%	\item Why is it hard: the computational complexity of searching
+%	  combinations of multiple modes that can be combined freely is high
+%	  (PSPACE-complete\cite{}) which is why existing trip planner exploit problem
+%	  structure (such as only one mode per trip) and predefined
+%	  preferences (optimization metrics)
+%	\item How did we solve it: Using a high-performance graph search
+%	  engine\cite{}, we implemented a multi-modal trip planner that uses pure
+%	  graph-search. This allows us to flexibly combine various modes
+%	  (e.g., car, train, cab, walk) and declarative change soft and hard
+%	  constraints. The planner can take constraints expressed in linear
+%	  temporal logic to constrain the search space (e.g., never bike after
+%	  transit, i.e., I don't want to take my bike on the bus or train).
+%	  The planner uses a weighted sum over several metrics (e.g., time
+%	  overall, time spend on transit, cost) which can be re-weighted based
+%	  on different user preferences. Finally the planner allows the user
+%	  to upload {\em new} mapping data over which to express hard and/or
+%	  soft constraints. For instance, a user might upload a map of crime
+%	  in the city, and ask the trip planner to never walk at night through
+%	  an area where crime is frequent.
+%%	\item How did we evaluate the features and what are the results:
+%%	  ceteris paribus experimentation setup on Amazon Mechanical 
+%%		Turk (MTurk\footnote{\url{http://mturk.amazon.com}}) and results, 
+%%		asking users to pick between two different trips, one being limited by the
+%%	  capabilities of existing planners, and one exhibiting features only
+%%	  available in our trip planner.
+%\end{itemize}
 
 \section{Extensibility}
-\begin{itemize}
-	\setlength\itemsep{1pt}
-	\item Explain in detail what extensibility means.
-	\item Describe the input.
-	\item Describe how we incorporate the auxiliary data into the planner.
-\end{itemize}
+%\begin{itemize}
+%	\setlength\itemsep{1pt}
+%	\item Explain in detail what extensibility means.
+%	\item Describe the input.
+%	\item Describe how we incorporate the auxiliary data into the planner.
+%\end{itemize}
 Allowing users to upload their own data sets of interests is an important
 step towards customization of a trip planner.
-We designed a framework where a user can load the data sets (e.g., crime
+We designed a framework where a user can upload auxiliary cost 
+metric data (e.g., crime
 statistics and pollution data) into the planner,
@@
 
-\subsection{Auxiliary Data}
 The user-created data are the auxiliary data that is represented as pairs
@@
 \subsection{Constraints}
-\begin{itemize}
-	\setlength\itemsep{1pt}
-	\item Describe syntax and semantics of LTL in the setting of trip planning.
-	\begin{itemize}
-		\setlength\itemsep{0pt}
-		\item Specifically, we need to point out the ``after" temporal connective,
-					and we need to describe the semantics: what is a plan, why we can
-					eliminate actions and put them as state attributes instead, etc.
-	\end{itemize}
-\end{itemize}
+%\begin{itemize}
+%	\setlength\itemsep{1pt}
+%	\item Describe syntax and semantics of LTL in the setting of trip planning.
+%	\begin{itemize}
+%		\setlength\itemsep{0pt}
+%		\item Specifically, we need to point out the ``after" temporal connective,
+%					and we need to describe the semantics: what is a plan, why we can
+%					eliminate actions and put them as state attributes instead, etc.
+%	\end{itemize}
+%\end{itemize}
 As constraints in the setting of trip planning are often declarative and
@@
   \centering
-    \includegraphics[width=0.5\textwidth]{figs/trajectories.pdf}
+    \includegraphics[width=0.5\textwidth]{figs/result1.png}
   \caption{State transition diagram\label{fig:trjs}}
@@
 \subsection{Preferences}
-\begin{itemize}
-	\setlength\itemsep{1pt}
-	\item Describe our preferential cost function (PCF).
-	\begin{itemize}
-		\setlength\itemsep{0pt}
-		\item Describe why and how we break down time and money.
-		\item Describe why and how we design coefficients to these time and money pieces.
-		\item Describe why and how we design the two ratios: dollars per hour and dollars per aux.
-	\end{itemize}
-\end{itemize}
+%\begin{itemize}
+%	\setlength\itemsep{1pt}
+%	\item Describe our preferential cost function (PCF).
+%	\begin{itemize}
+%		\setlength\itemsep{0pt}
+%		\item Describe why and how we break down time and money.
+%		\item Describe why and how we design coefficients to these time and money pieces.
+%		\item Describe why and how we design the two ratios: dollars per hour and dollars per aux.
+%	\end{itemize}
+%\end{itemize}
 

### 2016-03-02 paper.tex
Added figures. Need to change Figure 3 though.

@@
 \newcommand{\cA}{\mathcal{A}}
+\newcommand{\cU}{\mathcal{U}}
+\newcommand{\cP}{\mathcal{P}}
 \newcommand{\tit}[1]{\textit{#1}}
@@
 
-Formally, let $A$ be the auxiliary set of lat-long pairs, 
-$N$ a node in our planner and $r$ the effective radius.
-We first compute the set $P$ of auxiliary points such that
-$P \subseteq A$ and $\dist(p,N)<r$ for all $p \in P$.
-Then, the total score of auxiliary data for $N$ with respect to $P$ and $r$,
-denoted by $S_\aux(N,P,r)$, is computed as follows.
+Formally, let $\cU$ be the uploaded auxiliary dataset of lat-long pairs, 
+$N=(x_N,y_N)$ a node in our planner and $r$ the effective radius.
+We first compute the set $\cP \subseteq A$ of auxiliary points such that,
+for every node $P \in \cP$, the Euclidean distance between $N$ and $P$
+is within $r$.
+Then, the score of auxiliary data for $N$ with respect to $\cP$ and $r$,
+denoted by $S(N,\cP,r)$, is computed as follows.
 
-\begin{equation}
-	S_\aux(N,P,r) = \sum_{\substack{p \in P}} \Big(1 - \frac{\dist(p,N)}{r}\Big)
-\end{equation}
+\begin{equation*}
+	S(N,\cP,r) = \sum_{\substack{P \in \cP}} 
+		\Big(1 - \frac{\sqrt{(x_N-x_P)^2+(y_N-y_P)^2}}{r}\Big)
+\end{equation*}
+
+Now we turn to \figref{aux} for an instance to show how auxiliary data
+are integrated into the graph using the equation above.
+In \figref{aux}, we have the green nodes denoting the graph nodes, 
+and the red nodes the new auxiliary nodes that represent locations
+of criminal events.
+For node $N$, there are three red nodes within its neighborhood of
+radius of $r$; thus, we have $\cP=\{P_1,P_2,P_3\}$.
+We assume $r$ is 100 feet, and the distances from $N$ to $P_1$,
+$P_2$ and $P_3$ are 25, 80 and 90 feet, respectively.
+Then, the crime score of $S(N,\cP,r)$ is $1.05$.
+
+\begin{figure}[!ht]
+  \centering
+    \includegraphics[width=0.4\textwidth]{figs/aux.pdf}
+  \caption{Integrating uploaded data\label{fig:aux}}
+\end{figure}
 
@@
 
-An intuitive constraint in trip planning could be ``I will never drive after biking or
-taking the public transit."
+A natural constraint in trip planning could be ``In this trip I will not drive a car 
+after biking or taking the public transit."
 In LTL, such constraint can be translated into an LTL formula
@@
 \subsection{Reasoning with Constraints and Preferences}
-\begin{itemize}
-	\setlength\itemsep{1pt}
-	\item Describe why and how we integrate constraints and preferences into A*.
-\end{itemize}
+%\begin{itemize}
+%	\setlength\itemsep{1pt}
+%	\item Describe why and how we integrate constraints and preferences into A*.
+%\end{itemize}
+
+\begin{figure}[!ht]
+  \centering
+    \includegraphics[width=0.4\textwidth]{figs/Astar.pdf}
+  \caption{Adjusted A*\label{fig:astar}}
+\end{figure}
 
 We leveraged the widely-used A* search algorithm on top of our high-performance graph search
-engine.  The A* algorithm incorporates the following cost function.
+engine (cf. \figref{astar}).  The A* algorithm incorporates the following cost function.
 \begin{equation}
@@
 %\end{itemize}
-\section{Implementation}
+\section{Implementation and Results}
+%\begin{itemize}
+%	\setlength\itemsep{1pt}
+%	\item Describe the system structure.
+%	\item Describe the input to the system.
+%	\item Describe the output from the system.
+%\end{itemize}
 
-\begin{itemize}
-	\setlength\itemsep{1pt}
-	\item Describe the system structure.
-	\item Describe the input to the system.
-	\item Describe the output from the system.
-\end{itemize}
+\subsection{System Architecture}
+\begin{figure*}[!ht]
+  \centering
+    \includegraphics[width=0.8\textwidth]{figs/system.pdf}
+  \caption{System Overview\label{fig:system}}
+\end{figure*}
+
+\subsection{Results}
+First, we show resulting route for Alice in \figref{alice}.
+\begin{figure}[!ht]
+  \centering
+    \includegraphics[width=0.5\textwidth]{figs/result_Alice.pdf}
+  \caption{Resulting route for Alice\label{fig:alice}}
+\end{figure}
+
+Second, we show resulting route for Bob in \figref{bob}.
+\begin{figure}[!ht]
+  \centering
+    \includegraphics[width=0.5\textwidth]{figs/result_Bob.pdf}
+  \caption{Resulting route for Bob\label{fig:bob}}
+\end{figure}
+
+Third, we show resulting route for Cal in \figref{cal}.
+\begin{figure}[!ht]
+  \centering
+    \includegraphics[width=0.5\textwidth]{figs/result_Cal.pdf}
+  \caption{Resulting route for Cal\label{fig:cal}}
+\end{figure}
 

### 2016-03-02 paper.tex
Small save.
Updated section Preferences.

@@
 \newcommand{\corref}[1]{Corollary~\ref{cor:#1}}
+\newcommand{\eqtref}[1]{Equation~\ref{eq:#1}}
 
@@
 \title{On Extensibility and Personalizability of Multi-Modal Trip Planning}
-\author{
-	Xudong Liu\\
-	Department of Computer Science\\
-	University of Kentucky\\
-	329 Rose Street\\
-	Lexington, KY 40506\\
-	USA
-	\And
-	PARC People\\
-	Palo Alto Research Center\\
-	3333 Coyote Hill Road\\
-	Palo Alto, CA 94304\\
-	USA
-}
+%\author{
+%	Xudong Liu\\
+%	Department of Computer Science\\
+%	University of Kentucky\\
+%	329 Rose Street\\
+%	Lexington, KY 40506\\
+%	USA
+%	\And
+%	PARC People\\
+%	Palo Alto Research Center\\
+%	3333 Coyote Hill Road\\
+%	Palo Alto, CA 94304\\
+%	USA
+%}
 \maketitle
@@
   \centering
-    \includegraphics[width=0.5\textwidth]{figs/result1.png}
+    \includegraphics[width=0.4\textwidth]{figs/trajectories.pdf}
   \caption{State transition diagram\label{fig:trjs}}
@@
 The state variables of a state $S$ include the transportation mode $M$ that led to $S$,
-time spent $T$ so far per mode (e.g., $T_\bike$ for biking and $T_\public$ for
+time $T$ spent so far per mode (e.g., $T_\bike$ for biking and $T_\public$ for
 public transit), fare $D$ spent so far per mode (e.g., $D_\gas$ for driving and
@@
 	\begin{aligned}
-		\PCF(S) =& \beta_1 * (\alpha_1 \cdot T_\walk + \alpha_2 \cdot T_\wait + \ldots) \\
+		\PCF(S) =& \beta_1 * (\alpha_\walk \cdot T_\walk + \alpha_\bike \cdot T_\bike + \ldots) \\
 								&+ (D_\gas + D_\public + \ldots) \\
@@
 	\end{aligned}
+	\label{eq:pcf}
 \end{equation}
@@
 \subsubsection{Preference Elicitation}
-To gather these coefficients ($\alpha_i$ and $\beta_i$) in our $\PCF$, we designed interface to
+To gather these coefficients ($\alpha_i$'s and $\beta_i$'s) in our $\PCF$, we designed interface to
 elicit these numbers from the user.
+It asks the user questions and collect answers from the user to derive the coefficients.
+These questions are as follows.
+\begin{enumerate}
+	\item One hr of walking $=$ \underline{\hspace{1cm}} hrs of driving?
+	\item One hr of biking $=$ \underline{\hspace{1cm}} hrs of driving?
+	\item One hr of public transit $=$ \underline{\hspace{1cm}} hrs of driving?
+	\item One hr of taxi $=$ \underline{\hspace{1cm}} hrs of driving?
+	\item How much in dollars would you pay to save an hour in traveling?
+	\item How much in dollars would you pay to avoid an auxiliary event (e.g., crime) in traveling?
+\end{enumerate}
+For instance, the user answers 3, 2, 0.25 and 0.5 to the first four questions in the list above.
+Intuitively, the numbers indicate that the user prefers public transit the most, followed by
+taxi, driving, biking and walking, in order.
+We show how we can now derive $\alpha_i$'s in \eqtref{pcf}.
+We start with setting $\alpha_\car=1$.
+Now, since one hour of walking is equivalent to 3 hours of walking,
+we have $\alpha_\walk \times 1 = \alpha_\car \times 3$;
+hence, we derive $\alpha_\walk = 3$.
+Similarly, we have $\alpha_\bike = 2$, $\alpha_\public = 0.25$,
+and $\alpha_\taxi = 0.5$.
+Computing $\beta_1$ and $\beta_1$ is straightforward.
+Indeed, function $\PCF$ with the input of time, fare and
+auxiliary metric pieces boils down to monetary cost,
+and the planner computes the best path by optimize
+based on this overall monetary cost in the searching
+process.
 

### 2016-03-02 paper.tex
Updated sections Implementation and Results, and Conclusion and Future Work.

@@
 \usepackage{amsfonts}
+\usepackage{framed}
 
@@
 \newcommand{\corref}[1]{Corollary~\ref{cor:#1}}
-\newcommand{\eqtref}[1]{Equation~\ref{eq:#1}}
+\newcommand{\eqtref}[1]{Equation~\ref{eqt:#1}}
 
@@
 on different user preferences.
+
+Our paper is organized as follows.
+In the next section, we present what it means for a planner to be extensible
+and formally define the method to incorporating new metrics into the planner.
+In the following section, we discuss the two aspects of personalization
+in trip planning: constraints and preferences, and how they are modeled and
+reasoned with in the setting of multi-modal trip planning.
+We then move on to describe the system structure of our graph-search based planner,
+and show results obtained from our planner in various occassions.
+Finally, we conclude by outlining some future research directions.
 
@@
 In LTL, such constraint can be translated into an LTL formula
-\begin{align*}
+\begin{equation}
+\label{eqt:ex}
 	\psi = ((M=b) \lor (M=p)) \,\cA\, (\Box (\neg (M=c))).
-\end{align*}
+\end{equation}
 
@@
 	\end{aligned}
-	\label{eq:pcf}
+	\label{eqt:pcf}
 \end{equation}
@@
 \end{enumerate}
-For instance, the user answers 3, 2, 0.25 and 0.5 to the first four questions in the list above.
-Intuitively, the numbers indicate that the user prefers public transit the most, followed by
+For instance, Alice, an agent, answers 3, 2, 0.25 and 0.5 to the first four questions in the list above.
+Intuitively, the numbers indicate that she prefers public transit the most, followed by
 taxi, driving, biking and walking, in order.
@@
 \end{figure*}
+We designed and implemented a multi-modal trip planning
+system based on a high-performance graph search engine.
+The planner allows user uploads, as well as declarative
+constraints and preferences.
+We now describe the structure of the planning system.
+
+The trip planner takes two types of data as input:
+static data and user-specified request.
+The static input includes Map Data and Transit
+Data.
+Map Data describes the map, a directed graph where 
+nodes are street corners, bus stops and train stations.
+Transit Data is a set of schedules for the buses and trains
+On the other hand, a user provides her request, composed of 
+three parts.
+First, the user enters \tit{from} and \tit{to} locations 
+on the map together with day and time of the start of the trip.  
+Second, the user may upload her auxiliary metric dataset, e.g., crime rates.  
+Lastly, the user specifies her constraints in LTL and preferences as a $\PCF$.
+For example, the constraint 
+could be ``never walk through a bad neighborhood."  Given these 
+inputs, our planner computes an optimal path satisfying 
+all the constraints and optimizing the preferences.
+Note that the request from the user is encapsulated into
+a JSON object.
+
+For instance, the JSON object for the constraint $\psi$ in \eqtref{ex}
+is shown in \figref{exJSON}.
+\begin{figure}[!ht]
+  \centering
+    \includegraphics[width=0.5\textwidth]{figs/exampleJSON.png}
+  \caption{JSON object for constraint $\psi$ in \eqtref{ex}\label{fig:exJSON}}
+\end{figure}
 
 \subsection{Results}
-First, we show resulting route for Alice in \figref{alice}.
+We present and analyze resulting routes for three agents:
+Alice, Bob, and Cal.
+This is assuming no auxiliary metric datasets uploaded
+so that the agents focus on time and fare.
+Fixing their where and when information, we show how
+agents' different constraints and preferences affect
+their optimal routes computed by our planner.
+Their where and when are set so that they all plan to
+travel from San Jose International Airport (SJC) in
+San Jose, to Pier 39 in San Francisco.
+Note that the natural constraint $\psi$ in \eqtref{ex} 
+is implicitly imposed on all cases.
+
+Agent Alice has only one constraint that she does not 
+have a car; thus, driving to her is never available during this trip.
+Per her preferences, Alice provides her thoughts as earlier;
+that is, public transportation is preferred to taxi, which is
+better than biking, preferred to walking.
+What's more, she decides that she would sacrifice thirty dollars
+to save one hour during the trip.
+
+The resulting route for Alice is included in \figref{alice}.
+The solution takes 2 hours 7 minutes and 18 dollars 94 cents.
+It is optimal in the sense the combined value of time
+and fare is minimal among all paths from SJC to Pier 39.
+
 \begin{figure}[!ht]
@@
 
-Second, we show resulting route for Bob in \figref{bob}.
+Like Alice, Bob is constrained that he will not drive a car in his travel,
+and his dollar per hour is thirty.
+Moreover, Bob makes his mind to have some workout with his bike and
+dictates that he will bike for between one and two hours.
+He expresses his preferences: biking and public transit are the most
+preferred, next is taxi, and the least preferred is walking.
+Similarly, he has done so by answering the aforementioned elicitation questions.
+
+The result for Bob is depicted in \figref{bob}.
+It spans 2 hours 57 minutes in time with the fare of 29 dollars 17 cents.
+It is so, seemingly worse than what Alice achieved, only because Bob has
+the constraint that he will for sure bike for one to two hours, and
+the preferences that put biking the most satisfying mode.
+
 \begin{figure}[!ht]
@@
 
-Third, we show resulting route for Cal in \figref{cal}.
+Agent Cal will also not use a car in the trip.
+He affirms that time is greatly valuable and one hour is measured
+to 500 dollars, although his budget is restricted to 50.
+Cal's preferences are that the most preferred are public transit
+and taxi, and the next preferred are walking and biking that
+are equivalent.
+
+Refer to \figref{cal} for the optimal path for Cal.
+This route, stretching 1 hour 48 minutes, is the most time-saving one 
+compared with the previous two, at the price of 49 dollars 91 cents.
+This is due to the constraint that Cal can spend up to 50 dollars,
+as well as his preferences and vale of time being high.
+
 \begin{figure}[!ht]
@@
 
+\subsubsection{Auxiliary Metric}
+When the user of the planner is interested in metrics other than
+the ones offered already (i.e., time and fare), she might discover
+new metrics (e.g., crime rates and pollution statistics), upload
+them into the planner, and retrieve optimal plans taking these
+metrics into account.  One scenario of this approach is the following.
+
+Our agent needs to commute back home at night.
+
+\begin{figure}[!ht]
+  \centering
+    \includegraphics[width=0.5\textwidth]{figs/crime.pdf}
+  \caption{Optimal route considering crime rates\label{fig:crime}}
+\end{figure}
+
 \section{Conclusion and Future Work}
+In multi-modal trip planning, it is vital for the planner to enable
+users to integrate new metrics of interest into the existing map,
+and to express constraints and preferences over which optimal
+routes are computed.
+In our work, we designed and built a graph-search based, extensible
+and personalizable multi-modal trip planner that utilizes linear
+temporal logic and preferential cost
+function to model user constraints and preferences.
+
+As a note on the future work, we plan to explore techniques in planning
+for computing multiple routes that are diverse with bounded difference of costs.
+Continuing this direction, we intend to study the problem of learning the
+$\PCF$ coefficients using the observations of the decisions the user made among
+the computed paths. (It is reasonable to assume that routes picked by the user
+are preferred to others, and these observations could contribute to the learning
+problem.)
+Also interesting is to introduce traffic information into the planner
+to support real-time multi-agent concurrent trip planning.
 

### 2016-03-03 paper.tex
Spell check.

@@
 Trip planning, as an application of planning and scheduling,
-has seen substansive implementations by researchers and
+has seen substantive implementations by researchers and
 developers\cite{bast2015route}.
@@
 In this case, the planner will need to accommodate user preferences
-whenever possilbe in the search of optimal solutions.
+whenever possible in the search of optimal solutions.
 
@@
 We then move on to describe the system structure of our graph-search based planner,
-and show results obtained from our planner in various occassions.
+and show results obtained from our planner in various occasions.
 Finally, we conclude by outlining some future research directions.

### 2016-03-03 paper.tex
Updated section Results.

@@
 around midnight may want to upload a data set of crimes (cf. \figref{sf_crime}), 
-and express her constraints and preferences in hope of a preferred trip plan.
+and express her constraints and preferences in hope of a safer trip plan.
+Note that in \figref{sf_crime} bad areas are the colored circles, whose
+integer labels represent numbers of crimes in corresponding areas.
 
@@
 We designed and implemented a multi-modal trip planning
-system based on a high-performance graph search engine.
+system (cf. \figref{system}) based on a 
+high-performance graph search engine.
 The planner allows user uploads, as well as declarative
@@
 
-Our agent needs to commute back home at night.
+Our agent needs to travel without a car across San Francisco downtown at night.
+For her, safety is important.
+Having found the crime statistics for the area, the agent uploads
+the data as a new auxiliary metric into the map.
+By specifying that she will never walk through a neighborhood
+with more than fifteen crimes over the last month, and that
+she would sacrifice a quarter to avoid one crime incident,
+the agent tells the planner to come up with a relatively safe
+route. An example is shown in \figref{crime}, where the agent
+needs to start at the east of downtown and travel across the area
+to arrive at the west side.  
+The computed path is represented by the line colored by black,
+blue and green, denoting taxi, public transit and biking, respectively.
+Clearly, this path routes away from crime-heavy areas and achieves
+optimality in that the combined metrics -- time, fare and crime rates,
+uploaded and personalized by the user -- is minimal among all possibilities.
 

### 2016-03-03 paper.tex
Added reference paper on scenic routing.

@@
 such as crime rates and pollution statistics.
+One notable work by Nina et al.\cite{nina2016no} introduced
+system \tit{Autobahn} for generating scenic routes using
+Google Street View images to train a deep neural network
+to classify route segments based on their visual features.
+Although \tit{Autobahn} computes scenic routes using computer
+vision techniques, it does not account for extensibility
+and personalizability.
 

### 2016-03-03 extensibility.tex
Broke paper.tex in pieces.

--- 2016-03-03 paper.tex
@@
-\def\year{2016}\relax
-%File: formatting-instruction.tex
-\documentclass[letterpaper]{article}
-\usepackage{aaai16}
-\usepackage{times}
-\usepackage{helvet}
-\usepackage{courier}
-
-\usepackage[pdftex]{hyperref}
-\usepackage{amsmath}
-\usepackage{color}
-\usepackage{textcomp}
-\usepackage{booktabs}
-\usepackage{ccicons}
-\usepackage{todonotes}
-\usepackage{amsfonts}
-\usepackage{framed}
-
-\frenchspacing
-\setlength{\pdfpagewidth}{8.5in}
-\setlength{\pdfpageheight}{11in}
-\pdfinfo{
-/Title (On Extensibility and Personalizability of Multi-Modal Trip Planning)
-/Author (Put All Your Authors Here, Separated by Commas)}
-\setcounter{secnumdepth}{0}
-
-\newcommand{\aux}{\mathit{aux}}
-\newcommand{\dist}{\mathit{dist}}
-\newcommand{\cA}{\mathcal{A}}
-\newcommand{\cU}{\mathcal{U}}
-\newcommand{\cP}{\mathcal{P}}
-\newcommand{\tit}[1]{\textit{#1}}
-\newcommand{\IFF}{\textit{iff}}
-\newcommand{\IF}{\textit{if}}
-\newcommand{\PCF}{\mathit{PCF}}
-\newcommand{\walk}{\mathit{walk}}
-\newcommand{\bike}{\mathit{bike}}
-\newcommand{\car}{\mathit{car}}
-\newcommand{\public}{\mathit{public}}
-\newcommand{\taxi}{\mathit{taxi}}
-\newcommand{\gas}{\mathit{gas}}
-\newcommand{\wait}{\mathit{wait}}
-\newcommand{\SUM}{\mathit{sum}}
-\newcommand{\MAX}{\mathit{max}}
-\newcommand{\MIN}{\mathit{min}}
-\newcommand{\AVG}{\mathit{avg}}
-
-\newcommand{\secref}[1]{Section~\ref{sec:#1}}
-\newcommand{\figref}[1]{Figure~\ref{fig:#1}}
-\newcommand{\tblref}[1]{Table~\ref{tbl:#1}}
-\newcommand{\algref}[1]{Algorithm~\ref{alg:#1}}
-\newcommand{\thmref}[1]{Theorem~\ref{thm:#1}}
-\newcommand{\propref}[1]{Proposition~\ref{prop:#1}}
-\newcommand{\corref}[1]{Corollary~\ref{cor:#1}}
-\newcommand{\eqtref}[1]{Equation~\ref{eqt:#1}}
-
- \begin{document}
-% The file aaai.sty is the style file for AAAI Press 
-% proceedings, working notes, and technical reports.
-%
-\title{On Extensibility and Personalizability of Multi-Modal Trip Planning}
-%\author{
-%	Xudong Liu\\
-%	Department of Computer Science\\
-%	University of Kentucky\\
-%	329 Rose Street\\
-%	Lexington, KY 40506\\
-%	USA
-%	\And
-%	PARC People\\
-%	Palo Alto Research Center\\
-%	3333 Coyote Hill Road\\
-%	Palo Alto, CA 94304\\
-%	USA
-%}
-\maketitle
-\begin{abstract}
-	Trip planning is a practical task that has drawn
-	extensive attention from the AI planning and scheduling community
-	and the industrial/commercial sectors.
-	In this paper, we consider the setting of the \tit{multi-modal}
-	trip planning, where users can exploit different transportation
-	modes, such as walking, biking, public transit, and taxi.
-	In such a context, it would be of benefit if the
-	user was able to extend the \tit{cost base}, including traveling time 
-	and fare, of the planner, and to personalize the planner according
-	to her own constraints and preferences.
-	To this end, we designed and developed a hipergraph-based multi-modal
-	trip planner that allows users to upload auxiliary cost metrics (e.g.,
-	crime rates), to specify constraints as a theory in the 
-	\tit{linear temporal logic}, and to express preferences as a 
-	\tit{preferential cost function}.
-\end{abstract}
-
-\section{Introduction}
-Trip planning, as an application of planning and scheduling,
-has seen substantive implementations by researchers and
-developers\cite{bast2015route}.
-Some of the planning systems are \tit{multi-modal}; that is,
-combining distinct transportation modes, the trip planners
-compute optimal routes from sources to destinations.
-This notion of ``optimality" generally refers to the computed routes
-having minimal total time or total fare.
-
-However, in the eyes of a user, it may be more faceted than just
-``fastest" or ``cheapest."
-For instance, for a college student who specifies that she will
-only walk or take public transit in a trip from Palo Alto to
-San Francisco, the computed plan is not necessarily the fastest
-(e.g., taking a cab could be faster)
-or the cheapest (e.g., walking all the way has no fare).
-This happens when a user tells the planner her hard constraints,
-called \tit{constraints}.  
-The planner then needs to either satisfy the 
-constraints in the search process or return failure because they are
-over-restrictive.
-Moreover, the user might want to further customize the planner by
-describing soft constraints, called \tit{preferences}.
-For example, an agent wants to travel from school to
-downtown, and prefers biking to taking the bus.
-Thus, a trip with more biking than bus may be considered better for
-the agent than the one with more bus than biking.
-In this case, the planner will need to accommodate user preferences
-whenever possible in the search of optimal solutions.
-
-Representing and reasoning about constraints and preferences are
-fundamental to decision making in automated planning and scheduling
-in artificial intelligence.
-Son and Pontelli presented a declarative preference language, 
-$\mathcal{PP}$, to represent ordinal preferences (e.g., basic desires,
-atomic preferences and general preferences) between trajectories of 
-a planning problem\cite{son2004planning}.
-Bienvenu et al. proposed a first-order preference language, $\mathcal{LPP}$, 
-extending $\mathcal{PP}$ by allowing the specification of quantified qualitative
-preferences\cite{bienvenu2011specifying}.
-Modesti and Sciomachen introduced an ad hoc utility function for weighing
-the arcs both with their time and fare\cite{modesti1998utility}.
-This relationship between time and fare in the setting of transportation
-was studied, and economic models capturing the trade-off between the two
-was presented\cite{antoniou2007methodology}.
-
-However, relatively limited effort has been devoted to designing and
-implementing real-world multi-modal trip planners that captures user
-constraints and preferences over the \tit{cost base}, possibly
-extended from the user with auxiliary \tit{cost metrics},
-such as crime rates and pollution statistics.
-
-Using a high-performance graph search
-engine\cite{zhou2011dynamic}, we designed and implemented a 
-multi-modal trip planner that uses pure
-graph-search. This allows us to flexibly combine various modes
-(e.g., walk, bike, car, public transit, and taxi) and 
-declaratively change constraints and preferences. 
-The planner allows the user
-to upload \tit{new} mapping data over which to express
-constraints and preferences.  For instance, a user might upload a map of crime
-in the city, and ask the trip planner to avoid areas where crime is frequent.
-Then, the planner takes constraints expressed in \tit{linear
-temporal logic} to constrain the search space (e.g., never bike after
-transit, and never walk through bad neighborhoods).
-Finally, the planner uses a weighted sum, called \tit{preferential cost function}, 
-over several cost metrics (e.g., time spend biking, fare on public transit, 
-and overall crimes walking through) which can be re-weighted based
-on different user preferences.
-
-Our paper is organized as follows.
-In the next section, we present what it means for a planner to be extensible
-and formally define the method to incorporating new metrics into the planner.
-In the following section, we discuss the two aspects of personalization
-in trip planning: constraints and preferences, and how they are modeled and
-reasoned with in the setting of multi-modal trip planning.
-We then move on to describe the system structure of our graph-search based planner,
-and show results obtained from our planner in various occasions.
-Finally, we conclude by outlining some future research directions.
-
-%\begin{itemize}
-%	\item What's the problem: personalized trip planning: find trips that
-%	  are optimal in the eyes of a user, who doesn't care about mode or
-%	  provider boundaries, and whose preferences may be more faceted than
-%	  ``fastest'' or ``cheapest''.
-%	\item Why is it hard: the computational complexity of searching
-%	  combinations of multiple modes that can be combined freely is high
-%	  (PSPACE-complete\cite{}) which is why existing trip planner exploit problem
-%	  structure (such as only one mode per trip) and predefined
-%	  preferences (optimization metrics)
-%	\item How did we solve it: Using a high-performance graph search
-%	  engine\cite{}, we implemented a multi-modal trip planner that uses pure
-%	  graph-search. This allows us to flexibly combine various modes
-%	  (e.g., car, train, cab, walk) and declarative change soft and hard
-%	  constraints. The planner can take constraints expressed in linear
-%	  temporal logic to constrain the search space (e.g., never bike after
-%	  transit, i.e., I don't want to take my bike on the bus or train).
-%	  The planner uses a weighted sum over several metrics (e.g., time
-%	  overall, time spend on transit, cost) which can be re-weighted based
-%	  on different user preferences. Finally the planner allows the user
-%	  to upload {\em new} mapping data over which to express hard and/or
-%	  soft constraints. For instance, a user might upload a map of crime
-%	  in the city, and ask the trip planner to never walk at night through
-%	  an area where crime is frequent.
-%%	\item How did we evaluate the features and what are the results:
-%%	  ceteris paribus experimentation setup on Amazon Mechanical 
-%%		Turk (MTurk\footnote{\url{http://mturk.amazon.com}}) and results, 
-%%		asking users to pick between two different trips, one being limited by the
-%%	  capabilities of existing planners, and one exhibiting features only
-%%	  available in our trip planner.
-%\end{itemize}
-
-\section{Extensibility}
 %\begin{itemize}
@@
 \end{figure}
-
-\section{Personalizability}
-Personalizability consists of two aspects: constraints and preferences.
-From the viewpoint of the planner,
-constraints, also referred to as hard constraints, are statements that the planner
-has to satisfy during the planning process; whereas preferences, also called
-soft constraints, are specifications that the planner will need to optimize.
-We formulated constraints using linear temporal logic (LTL) and preferences as
-a preferential cost function (PCF), and implemented our planner leveraging the
-widely-used graph search algorithm the A*.
-
-\subsection{Constraints}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Describe syntax and semantics of LTL in the setting of trip planning.
-%	\begin{itemize}
-%		\setlength\itemsep{0pt}
-%		\item Specifically, we need to point out the ``after" temporal connective,
-%					and we need to describe the semantics: what is a plan, why we can
-%					eliminate actions and put them as state attributes instead, etc.
-%	\end{itemize}
-%\end{itemize}
-As constraints in the setting of trip planning are often declarative and
-temporal, our choice of LTL is straightforward.
-We now give a brief review of linear temporal logic (LTL).
-Let $f$ be a propositional formula over a finite set $L$ of Boolean variables.  
-LTL formulas are defined recursively as follows.
-\begin{equation}
-	\varphi = f | \varphi_1 \land \varphi_2 | \varphi_1 \lor \varphi_2 | \neg \varphi | 
-		\bigcirc \varphi |	\Box \varphi | \Diamond \varphi | \varphi_1 \cA \varphi_2
-\end{equation}
-Note that we have $\varphi_1 \cA \varphi_2$, and it means that
-``$\varphi_2$ holds right after $\varphi_1$ holds."
-
-A natural constraint in trip planning could be ``In this trip I will not drive a car 
-after biking or taking the public transit."
-In LTL, such constraint can be translated into an LTL formula
-\begin{equation}
-\label{eqt:ex}
-	\psi = ((M=b) \lor (M=p)) \,\cA\, (\Box (\neg (M=c))).
-\end{equation}
-
-As the actions in trip planning is limited to taking different transportation modes,
-in our definition of the semantics of LTL
-these actions are subsumed into the interpretations of $L$, or \tit{states}.
-The semantics of LTL is defined with regard to trajectories of states. 
-Let $\sigma$ be a trajectory of states $S_0,a_1,S_1,\ldots,a_n,S_n$, and
-$\sigma[i]$ a suffix $S_i, a_{i+1}, S_{i+1}, \ldots,a_n,S_n$.  We have
-\begin{align*}
-	\sigma \models f \;\; &\IFF \;\; S_0 \models f,\\
-	\sigma \models \varphi_1 \land \varphi_2 \;\; &\IFF \;\; \sigma \models \varphi_1 \; and \; \sigma \models \varphi_2,\\
-	\sigma \models \varphi_1 \lor \varphi_2 \;\; &\IFF \;\; \sigma \models \varphi_1 \; or \; \sigma \models \varphi_2,\\
-	\sigma \models \neg \varphi \;\; &\IFF \;\; \sigma \not \models \varphi,\\
-	\sigma \models \bigcirc \varphi \;\; &\IFF \;\; \sigma[1] \models \varphi,\\
-	\sigma \models \Box \varphi \;\; &\IFF \;\; \forall 0 \leq i \leq n (\sigma[i] \models \varphi),\\
-	\sigma \models \Diamond \varphi \;\; &\IFF \;\; \exists 0 \leq i \leq n (\sigma[i] \models \varphi),\\
-	\sigma \models \varphi_1 \cA \varphi_2 \;\; &\IFF \;\; \forall 0 \leq i < n (\IF \; \sigma[i] \models \varphi_1, \sigma[i+1] \models \varphi_2).
-\end{align*}
-
-For example, we are given an LTL constraint $\psi$ and three trajectories ($\sigma_1$, $\sigma_2$ and $\sigma_3$)
-as shown in \figref{trjs}.
-Clearly, we have $\sigma_1 \models \psi$ because, after public transit in $S_2$ and $S_3$, 
-traveling by car has never been taken place.
-Moreover, we have $\sigma_2 \not \models \psi$ because we have $M=c$ hold in $S_6$ and $S_7$
-after having $M=p$ hold in $S_2$ and $S_5$.
-Finally, we know $\sigma_3 \models \psi$, as the mode is always neither biking nor public transit.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.4\textwidth]{figs/trajectories.pdf}
-  \caption{State transition diagram\label{fig:trjs}}
-\end{figure}
-
-\subsection{Preferences}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Describe our preferential cost function (PCF).
-%	\begin{itemize}
-%		\setlength\itemsep{0pt}
-%		\item Describe why and how we break down time and money.
-%		\item Describe why and how we design coefficients to these time and money pieces.
-%		\item Describe why and how we design the two ratios: dollars per hour and dollars per aux.
-%	\end{itemize}
-%\end{itemize}
-
-A state is described as a set of \tit{state variables}.
-The state variables of a state $S$ include the transportation mode $M$ that led to $S$,
-time $T$ spent so far per mode (e.g., $T_\bike$ for biking and $T_\public$ for
-public transit), fare $D$ spent so far per mode (e.g., $D_\gas$ for driving and
-$D_\taxi$ for taking a cab), and variables related to the auxiliary data once uploaded.
-These extra data related variables are metrics such as the sum ($A_\SUM$),
-the maximum ($A_\MAX)$, the minimum ($A_\MIN$), and the average ($A_\AVG$) data along the path.
-
-We focused on weighted functions over state variables and
-designed the cost function, called preferential cost function (PCF), that guides the
-graph-based search engine in our trip planner as follows.
-\begin{equation}
-	\begin{aligned}
-		\PCF(S) =& \beta_1 * (\alpha_\walk \cdot T_\walk + \alpha_\bike \cdot T_\bike + \ldots) \\
-								&+ (D_\gas + D_\public + \ldots) \\
-								&+ \beta_2 * (A_\SUM + \ldots),
-	\end{aligned}
-	\label{eqt:pcf}
-\end{equation}
-where $\alpha_i$ are real numbers representing the relations among different time pieces,
-and $\beta_1$ ($\beta_2$) is the ratio that essentially describes how much in dollars a user would pay to
-save an hour (an auxiliary data, respectively).
-
-\subsubsection{Preference Elicitation}
-To gather these coefficients ($\alpha_i$'s and $\beta_i$'s) in our $\PCF$, we designed interface to
-elicit these numbers from the user.
-It asks the user questions and collect answers from the user to derive the coefficients.
-These questions are as follows.
-\begin{enumerate}
-	\item One hr of walking $=$ \underline{\hspace{1cm}} hrs of driving?
-	\item One hr of biking $=$ \underline{\hspace{1cm}} hrs of driving?
-	\item One hr of public transit $=$ \underline{\hspace{1cm}} hrs of driving?
-	\item One hr of taxi $=$ \underline{\hspace{1cm}} hrs of driving?
-	\item How much in dollars would you pay to save an hour in traveling?
-	\item How much in dollars would you pay to avoid an auxiliary event (e.g., crime) in traveling?
-\end{enumerate}
-For instance, Alice, an agent, answers 3, 2, 0.25 and 0.5 to the first four questions in the list above.
-Intuitively, the numbers indicate that she prefers public transit the most, followed by
-taxi, driving, biking and walking, in order.
-We show how we can now derive $\alpha_i$'s in \eqtref{pcf}.
-We start with setting $\alpha_\car=1$.
-Now, since one hour of walking is equivalent to 3 hours of walking,
-we have $\alpha_\walk \times 1 = \alpha_\car \times 3$;
-hence, we derive $\alpha_\walk = 3$.
-Similarly, we have $\alpha_\bike = 2$, $\alpha_\public = 0.25$,
-and $\alpha_\taxi = 0.5$.
-Computing $\beta_1$ and $\beta_1$ is straightforward.
-Indeed, function $\PCF$ with the input of time, fare and
-auxiliary metric pieces boils down to monetary cost,
-and the planner computes the best path by optimize
-based on this overall monetary cost in the searching
-process.
-
-\subsection{Reasoning with Constraints and Preferences}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Describe why and how we integrate constraints and preferences into A*.
-%\end{itemize}
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.4\textwidth]{figs/Astar.pdf}
-  \caption{Adjusted A*\label{fig:astar}}
-\end{figure}
-
-We leveraged the widely-used A* search algorithm on top of our high-performance graph search
-engine (cf. \figref{astar}).  The A* algorithm incorporates the following cost function.
-\begin{equation}
-	f(S) = g(S) + h(S),
-\end{equation}
-where $g(S)$ is the cost of an optimal trip from the initial state to $S$, and
-$h(S)$ is an admissible estimate of the cost of an optimal trip from $S$ to goal.
-
-To prune the search space, we check satisfiability of the temporal constraints in LTL
-at expansion of the search tree.
-To guide the search engine, we set $g(S)=\PCF(S)$ and $h(S)$ the minimum estimate among
-all available modes in $S$.
-
-%\section{User Study}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Describe the set up of our experiments on AMT and explain why: because we want to evaluate the
-%				feasibility of the three features.
-%	\item Describe the results.
-%\end{itemize}
-\section{Implementation and Results}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Describe the system structure.
-%	\item Describe the input to the system.
-%	\item Describe the output from the system.
-%\end{itemize}
-
-\subsection{System Architecture}
-\begin{figure*}[!ht]
-  \centering
-    \includegraphics[width=0.8\textwidth]{figs/system.pdf}
-  \caption{System Overview\label{fig:system}}
-\end{figure*}
-We designed and implemented a multi-modal trip planning
-system based on a high-performance graph search engine.
-The planner allows user uploads, as well as declarative
-constraints and preferences.
-We now describe the structure of the planning system.
-
-The trip planner takes two types of data as input:
-static data and user-specified request.
-The static input includes Map Data and Transit
-Data.
-Map Data describes the map, a directed graph where 
-nodes are street corners, bus stops and train stations.
-Transit Data is a set of schedules for the buses and trains
-On the other hand, a user provides her request, composed of 
-three parts.
-First, the user enters \tit{from} and \tit{to} locations 
-on the map together with day and time of the start of the trip.  
-Second, the user may upload her auxiliary metric dataset, e.g., crime rates.  
-Lastly, the user specifies her constraints in LTL and preferences as a $\PCF$.
-For example, the constraint 
-could be ``never walk through a bad neighborhood."  Given these 
-inputs, our planner computes an optimal path satisfying 
-all the constraints and optimizing the preferences.
-Note that the request from the user is encapsulated into
-a JSON object.
-
-For instance, the JSON object for the constraint $\psi$ in \eqtref{ex}
-is shown in \figref{exJSON}.
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/exampleJSON.png}
-  \caption{JSON object for constraint $\psi$ in \eqtref{ex}\label{fig:exJSON}}
-\end{figure}
-
-\subsection{Results}
-We present and analyze resulting routes for three agents:
-Alice, Bob, and Cal.
-This is assuming no auxiliary metric datasets uploaded
-so that the agents focus on time and fare.
-Fixing their where and when information, we show how
-agents' different constraints and preferences affect
-their optimal routes computed by our planner.
-Their where and when are set so that they all plan to
-travel from San Jose International Airport (SJC) in
-San Jose, to Pier 39 in San Francisco.
-Note that the natural constraint $\psi$ in \eqtref{ex} 
-is implicitly imposed on all cases.
-
-Agent Alice has only one constraint that she does not 
-have a car; thus, driving to her is never available during this trip.
-Per her preferences, Alice provides her thoughts as earlier;
-that is, public transportation is preferred to taxi, which is
-better than biking, preferred to walking.
-What's more, she decides that she would sacrifice thirty dollars
-to save one hour during the trip.
-
-The resulting route for Alice is included in \figref{alice}.
-The solution takes 2 hours 7 minutes and 18 dollars 94 cents.
-It is optimal in the sense the combined value of time
-and fare is minimal among all paths from SJC to Pier 39.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/result_Alice.pdf}
-  \caption{Resulting route for Alice\label{fig:alice}}
-\end{figure}
-
-Like Alice, Bob is constrained that he will not drive a car in his travel,
-and his dollar per hour is thirty.
-Moreover, Bob makes his mind to have some workout with his bike and
-dictates that he will bike for between one and two hours.
-He expresses his preferences: biking and public transit are the most
-preferred, next is taxi, and the least preferred is walking.
-Similarly, he has done so by answering the aforementioned elicitation questions.
-
-The result for Bob is depicted in \figref{bob}.
-It spans 2 hours 57 minutes in time with the fare of 29 dollars 17 cents.
-It is so, seemingly worse than what Alice achieved, only because Bob has
-the constraint that he will for sure bike for one to two hours, and
-the preferences that put biking the most satisfying mode.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/result_Bob.pdf}
-  \caption{Resulting route for Bob\label{fig:bob}}
-\end{figure}
-
-Agent Cal will also not use a car in the trip.
-He affirms that time is greatly valuable and one hour is measured
-to 500 dollars, although his budget is restricted to 50.
-Cal's preferences are that the most preferred are public transit
-and taxi, and the next preferred are walking and biking that
-are equivalent.
-
-Refer to \figref{cal} for the optimal path for Cal.
-This route, stretching 1 hour 48 minutes, is the most time-saving one 
-compared with the previous two, at the price of 49 dollars 91 cents.
-This is due to the constraint that Cal can spend up to 50 dollars,
-as well as his preferences and vale of time being high.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/result_Cal.pdf}
-  \caption{Resulting route for Cal\label{fig:cal}}
-\end{figure}
-
-\subsubsection{Auxiliary Metric}
-When the user of the planner is interested in metrics other than
-the ones offered already (i.e., time and fare), she might discover
-new metrics (e.g., crime rates and pollution statistics), upload
-them into the planner, and retrieve optimal plans taking these
-metrics into account.  One scenario of this approach is the following.
-
-Our agent needs to commute back home at night.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/crime.pdf}
-  \caption{Optimal route considering crime rates\label{fig:crime}}
-\end{figure}
-
-\section{Conclusion and Future Work}
-In multi-modal trip planning, it is vital for the planner to enable
-users to integrate new metrics of interest into the existing map,
-and to express constraints and preferences over which optimal
-routes are computed.
-In our work, we designed and built a graph-search based, extensible
-and personalizable multi-modal trip planner that utilizes linear
-temporal logic and preferential cost
-function to model user constraints and preferences.
-
-As a note on the future work, we plan to explore techniques in planning
-for computing multiple routes that are diverse with bounded difference of costs.
-Continuing this direction, we intend to study the problem of learning the
-$\PCF$ coefficients using the observations of the decisions the user made among
-the computed paths. (It is reasonable to assume that routes picked by the user
-are preferred to others, and these observations could contribute to the learning
-problem.)
-Also interesting is to introduce traffic information into the planner
-to support real-time multi-agent concurrent trip planning.
-
-\bibliographystyle{aaai}
-\bibliography{refs}
-
-\end{document}

### 2016-03-03 implementation.tex
Broke paper.tex in pieces.

--- 2016-03-03 paper.tex
@@
-\def\year{2016}\relax
-%File: formatting-instruction.tex
-\documentclass[letterpaper]{article}
-\usepackage{aaai16}
-\usepackage{times}
-\usepackage{helvet}
-\usepackage{courier}
-
-\usepackage[pdftex]{hyperref}
-\usepackage{amsmath}
-\usepackage{color}
-\usepackage{textcomp}
-\usepackage{booktabs}
-\usepackage{ccicons}
-\usepackage{todonotes}
-\usepackage{amsfonts}
-\usepackage{framed}
-
-\frenchspacing
-\setlength{\pdfpagewidth}{8.5in}
-\setlength{\pdfpageheight}{11in}
-\pdfinfo{
-/Title (On Extensibility and Personalizability of Multi-Modal Trip Planning)
-/Author (Put All Your Authors Here, Separated by Commas)}
-\setcounter{secnumdepth}{0}
-
-\newcommand{\aux}{\mathit{aux}}
-\newcommand{\dist}{\mathit{dist}}
-\newcommand{\cA}{\mathcal{A}}
-\newcommand{\cU}{\mathcal{U}}
-\newcommand{\cP}{\mathcal{P}}
-\newcommand{\tit}[1]{\textit{#1}}
-\newcommand{\IFF}{\textit{iff}}
-\newcommand{\IF}{\textit{if}}
-\newcommand{\PCF}{\mathit{PCF}}
-\newcommand{\walk}{\mathit{walk}}
-\newcommand{\bike}{\mathit{bike}}
-\newcommand{\car}{\mathit{car}}
-\newcommand{\public}{\mathit{public}}
-\newcommand{\taxi}{\mathit{taxi}}
-\newcommand{\gas}{\mathit{gas}}
-\newcommand{\wait}{\mathit{wait}}
-\newcommand{\SUM}{\mathit{sum}}
-\newcommand{\MAX}{\mathit{max}}
-\newcommand{\MIN}{\mathit{min}}
-\newcommand{\AVG}{\mathit{avg}}
-
-\newcommand{\secref}[1]{Section~\ref{sec:#1}}
-\newcommand{\figref}[1]{Figure~\ref{fig:#1}}
-\newcommand{\tblref}[1]{Table~\ref{tbl:#1}}
-\newcommand{\algref}[1]{Algorithm~\ref{alg:#1}}
-\newcommand{\thmref}[1]{Theorem~\ref{thm:#1}}
-\newcommand{\propref}[1]{Proposition~\ref{prop:#1}}
-\newcommand{\corref}[1]{Corollary~\ref{cor:#1}}
-\newcommand{\eqtref}[1]{Equation~\ref{eqt:#1}}
-
- \begin{document}
-% The file aaai.sty is the style file for AAAI Press 
-% proceedings, working notes, and technical reports.
-%
-\title{On Extensibility and Personalizability of Multi-Modal Trip Planning}
-%\author{
-%	Xudong Liu\\
-%	Department of Computer Science\\
-%	University of Kentucky\\
-%	329 Rose Street\\
-%	Lexington, KY 40506\\
-%	USA
-%	\And
-%	PARC People\\
-%	Palo Alto Research Center\\
-%	3333 Coyote Hill Road\\
-%	Palo Alto, CA 94304\\
-%	USA
-%}
-\maketitle
-\begin{abstract}
-	Trip planning is a practical task that has drawn
-	extensive attention from the AI planning and scheduling community
-	and the industrial/commercial sectors.
-	In this paper, we consider the setting of the \tit{multi-modal}
-	trip planning, where users can exploit different transportation
-	modes, such as walking, biking, public transit, and taxi.
-	In such a context, it would be of benefit if the
-	user was able to extend the \tit{cost base}, including traveling time 
-	and fare, of the planner, and to personalize the planner according
-	to her own constraints and preferences.
-	To this end, we designed and developed a hipergraph-based multi-modal
-	trip planner that allows users to upload auxiliary cost metrics (e.g.,
-	crime rates), to specify constraints as a theory in the 
-	\tit{linear temporal logic}, and to express preferences as a 
-	\tit{preferential cost function}.
-\end{abstract}
-
-\section{Introduction}
-Trip planning, as an application of planning and scheduling,
-has seen substantive implementations by researchers and
-developers\cite{bast2015route}.
-Some of the planning systems are \tit{multi-modal}; that is,
-combining distinct transportation modes, the trip planners
-compute optimal routes from sources to destinations.
-This notion of ``optimality" generally refers to the computed routes
-having minimal total time or total fare.
-
-However, in the eyes of a user, it may be more faceted than just
-``fastest" or ``cheapest."
-For instance, for a college student who specifies that she will
-only walk or take public transit in a trip from Palo Alto to
-San Francisco, the computed plan is not necessarily the fastest
-(e.g., taking a cab could be faster)
-or the cheapest (e.g., walking all the way has no fare).
-This happens when a user tells the planner her hard constraints,
-called \tit{constraints}.  
-The planner then needs to either satisfy the 
-constraints in the search process or return failure because they are
-over-restrictive.
-Moreover, the user might want to further customize the planner by
-describing soft constraints, called \tit{preferences}.
-For example, an agent wants to travel from school to
-downtown, and prefers biking to taking the bus.
-Thus, a trip with more biking than bus may be considered better for
-the agent than the one with more bus than biking.
-In this case, the planner will need to accommodate user preferences
-whenever possible in the search of optimal solutions.
-
-Representing and reasoning about constraints and preferences are
-fundamental to decision making in automated planning and scheduling
-in artificial intelligence.
-Son and Pontelli presented a declarative preference language, 
-$\mathcal{PP}$, to represent ordinal preferences (e.g., basic desires,
-atomic preferences and general preferences) between trajectories of 
-a planning problem\cite{son2004planning}.
-Bienvenu et al. proposed a first-order preference language, $\mathcal{LPP}$, 
-extending $\mathcal{PP}$ by allowing the specification of quantified qualitative
-preferences\cite{bienvenu2011specifying}.
-Modesti and Sciomachen introduced an ad hoc utility function for weighing
-the arcs both with their time and fare\cite{modesti1998utility}.
-This relationship between time and fare in the setting of transportation
-was studied, and economic models capturing the trade-off between the two
-was presented\cite{antoniou2007methodology}.
-
-However, relatively limited effort has been devoted to designing and
-implementing real-world multi-modal trip planners that captures user
-constraints and preferences over the \tit{cost base}, possibly
-extended from the user with auxiliary \tit{cost metrics},
-such as crime rates and pollution statistics.
-
-Using a high-performance graph search
-engine\cite{zhou2011dynamic}, we designed and implemented a 
-multi-modal trip planner that uses pure
-graph-search. This allows us to flexibly combine various modes
-(e.g., walk, bike, car, public transit, and taxi) and 
-declaratively change constraints and preferences. 
-The planner allows the user
-to upload \tit{new} mapping data over which to express
-constraints and preferences.  For instance, a user might upload a map of crime
-in the city, and ask the trip planner to avoid areas where crime is frequent.
-Then, the planner takes constraints expressed in \tit{linear
-temporal logic} to constrain the search space (e.g., never bike after
-transit, and never walk through bad neighborhoods).
-Finally, the planner uses a weighted sum, called \tit{preferential cost function}, 
-over several cost metrics (e.g., time spend biking, fare on public transit, 
-and overall crimes walking through) which can be re-weighted based
-on different user preferences.
-
-Our paper is organized as follows.
-In the next section, we present what it means for a planner to be extensible
-and formally define the method to incorporating new metrics into the planner.
-In the following section, we discuss the two aspects of personalization
-in trip planning: constraints and preferences, and how they are modeled and
-reasoned with in the setting of multi-modal trip planning.
-We then move on to describe the system structure of our graph-search based planner,
-and show results obtained from our planner in various occasions.
-Finally, we conclude by outlining some future research directions.
-
-%\begin{itemize}
-%	\item What's the problem: personalized trip planning: find trips that
-%	  are optimal in the eyes of a user, who doesn't care about mode or
-%	  provider boundaries, and whose preferences may be more faceted than
-%	  ``fastest'' or ``cheapest''.
-%	\item Why is it hard: the computational complexity of searching
-%	  combinations of multiple modes that can be combined freely is high
-%	  (PSPACE-complete\cite{}) which is why existing trip planner exploit problem
-%	  structure (such as only one mode per trip) and predefined
-%	  preferences (optimization metrics)
-%	\item How did we solve it: Using a high-performance graph search
-%	  engine\cite{}, we implemented a multi-modal trip planner that uses pure
-%	  graph-search. This allows us to flexibly combine various modes
-%	  (e.g., car, train, cab, walk) and declarative change soft and hard
-%	  constraints. The planner can take constraints expressed in linear
-%	  temporal logic to constrain the search space (e.g., never bike after
-%	  transit, i.e., I don't want to take my bike on the bus or train).
-%	  The planner uses a weighted sum over several metrics (e.g., time
-%	  overall, time spend on transit, cost) which can be re-weighted based
-%	  on different user preferences. Finally the planner allows the user
-%	  to upload {\em new} mapping data over which to express hard and/or
-%	  soft constraints. For instance, a user might upload a map of crime
-%	  in the city, and ask the trip planner to never walk at night through
-%	  an area where crime is frequent.
-%%	\item How did we evaluate the features and what are the results:
-%%	  ceteris paribus experimentation setup on Amazon Mechanical 
-%%		Turk (MTurk\footnote{\url{http://mturk.amazon.com}}) and results, 
-%%		asking users to pick between two different trips, one being limited by the
-%%	  capabilities of existing planners, and one exhibiting features only
-%%	  available in our trip planner.
-%\end{itemize}
-
-\section{Extensibility}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Explain in detail what extensibility means.
-%	\item Describe the input.
-%	\item Describe how we incorporate the auxiliary data into the planner.
-%\end{itemize}
-Allowing users to upload their own data sets of interests is an important
-step towards customization of a trip planner.
-We designed a framework where a user can upload auxiliary cost 
-metric data (e.g., crime
-statistics and pollution data) into the planner,
-and the planner will compute an optimal route accordingly.
-
-The user-created data are the auxiliary data that is represented as pairs
-of latitude and longitude degrees.
-To merge these lat-long pairs into the planner, we performed a neighborhood
-search to calculate the total score of auxiliary data for each 
-lat-long pair already in our planner.
-It might be of strong interest to some user for our planning
-system to take care of criminal statistics so that some level of safety
-of the resulting routes is guaranteed.
-For instance, a user traveling through the downtown area of San Francisco
-around midnight may want to upload a data set of crimes (cf. \figref{sf_crime}), 
-and express her constraints and preferences in hope of a preferred trip plan.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.4\textwidth]{figs/sub_sf_crime_July2015.png}
-  \caption{Crime rates in San Francisco\label{fig:sf_crime}}
-\end{figure}
-
-Formally, let $\cU$ be the uploaded auxiliary dataset of lat-long pairs, 
-$N=(x_N,y_N)$ a node in our planner and $r$ the effective radius.
-We first compute the set $\cP \subseteq A$ of auxiliary points such that,
-for every node $P \in \cP$, the Euclidean distance between $N$ and $P$
-is within $r$.
-Then, the score of auxiliary data for $N$ with respect to $\cP$ and $r$,
-denoted by $S(N,\cP,r)$, is computed as follows.
-
-\begin{equation*}
-	S(N,\cP,r) = \sum_{\substack{P \in \cP}} 
-		\Big(1 - \frac{\sqrt{(x_N-x_P)^2+(y_N-y_P)^2}}{r}\Big)
-\end{equation*}
-
-Now we turn to \figref{aux} for an instance to show how auxiliary data
-are integrated into the graph using the equation above.
-In \figref{aux}, we have the green nodes denoting the graph nodes, 
-and the red nodes the new auxiliary nodes that represent locations
-of criminal events.
-For node $N$, there are three red nodes within its neighborhood of
-radius of $r$; thus, we have $\cP=\{P_1,P_2,P_3\}$.
-We assume $r$ is 100 feet, and the distances from $N$ to $P_1$,
-$P_2$ and $P_3$ are 25, 80 and 90 feet, respectively.
-Then, the crime score of $S(N,\cP,r)$ is $1.05$.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.4\textwidth]{figs/aux.pdf}
-  \caption{Integrating uploaded data\label{fig:aux}}
-\end{figure}
-
-\section{Personalizability}
-Personalizability consists of two aspects: constraints and preferences.
-From the viewpoint of the planner,
-constraints, also referred to as hard constraints, are statements that the planner
-has to satisfy during the planning process; whereas preferences, also called
-soft constraints, are specifications that the planner will need to optimize.
-We formulated constraints using linear temporal logic (LTL) and preferences as
-a preferential cost function (PCF), and implemented our planner leveraging the
-widely-used graph search algorithm the A*.
-
-\subsection{Constraints}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Describe syntax and semantics of LTL in the setting of trip planning.
-%	\begin{itemize}
-%		\setlength\itemsep{0pt}
-%		\item Specifically, we need to point out the ``after" temporal connective,
-%					and we need to describe the semantics: what is a plan, why we can
-%					eliminate actions and put them as state attributes instead, etc.
-%	\end{itemize}
-%\end{itemize}
-As constraints in the setting of trip planning are often declarative and
-temporal, our choice of LTL is straightforward.
-We now give a brief review of linear temporal logic (LTL).
-Let $f$ be a propositional formula over a finite set $L$ of Boolean variables.  
-LTL formulas are defined recursively as follows.
-\begin{equation}
-	\varphi = f | \varphi_1 \land \varphi_2 | \varphi_1 \lor \varphi_2 | \neg \varphi | 
-		\bigcirc \varphi |	\Box \varphi | \Diamond \varphi | \varphi_1 \cA \varphi_2
-\end{equation}
-Note that we have $\varphi_1 \cA \varphi_2$, and it means that
-``$\varphi_2$ holds right after $\varphi_1$ holds."
-
-A natural constraint in trip planning could be ``In this trip I will not drive a car 
-after biking or taking the public transit."
-In LTL, such constraint can be translated into an LTL formula
-\begin{equation}
-\label{eqt:ex}
-	\psi = ((M=b) \lor (M=p)) \,\cA\, (\Box (\neg (M=c))).
-\end{equation}
-
-As the actions in trip planning is limited to taking different transportation modes,
-in our definition of the semantics of LTL
-these actions are subsumed into the interpretations of $L$, or \tit{states}.
-The semantics of LTL is defined with regard to trajectories of states. 
-Let $\sigma$ be a trajectory of states $S_0,a_1,S_1,\ldots,a_n,S_n$, and
-$\sigma[i]$ a suffix $S_i, a_{i+1}, S_{i+1}, \ldots,a_n,S_n$.  We have
-\begin{align*}
-	\sigma \models f \;\; &\IFF \;\; S_0 \models f,\\
-	\sigma \models \varphi_1 \land \varphi_2 \;\; &\IFF \;\; \sigma \models \varphi_1 \; and \; \sigma \models \varphi_2,\\
-	\sigma \models \varphi_1 \lor \varphi_2 \;\; &\IFF \;\; \sigma \models \varphi_1 \; or \; \sigma \models \varphi_2,\\
-	\sigma \models \neg \varphi \;\; &\IFF \;\; \sigma \not \models \varphi,\\
-	\sigma \models \bigcirc \varphi \;\; &\IFF \;\; \sigma[1] \models \varphi,\\
-	\sigma \models \Box \varphi \;\; &\IFF \;\; \forall 0 \leq i \leq n (\sigma[i] \models \varphi),\\
-	\sigma \models \Diamond \varphi \;\; &\IFF \;\; \exists 0 \leq i \leq n (\sigma[i] \models \varphi),\\
-	\sigma \models \varphi_1 \cA \varphi_2 \;\; &\IFF \;\; \forall 0 \leq i < n (\IF \; \sigma[i] \models \varphi_1, \sigma[i+1] \models \varphi_2).
-\end{align*}
-
-For example, we are given an LTL constraint $\psi$ and three trajectories ($\sigma_1$, $\sigma_2$ and $\sigma_3$)
-as shown in \figref{trjs}.
-Clearly, we have $\sigma_1 \models \psi$ because, after public transit in $S_2$ and $S_3$, 
-traveling by car has never been taken place.
-Moreover, we have $\sigma_2 \not \models \psi$ because we have $M=c$ hold in $S_6$ and $S_7$
-after having $M=p$ hold in $S_2$ and $S_5$.
-Finally, we know $\sigma_3 \models \psi$, as the mode is always neither biking nor public transit.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.4\textwidth]{figs/trajectories.pdf}
-  \caption{State transition diagram\label{fig:trjs}}
-\end{figure}
-
-\subsection{Preferences}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Describe our preferential cost function (PCF).
-%	\begin{itemize}
-%		\setlength\itemsep{0pt}
-%		\item Describe why and how we break down time and money.
-%		\item Describe why and how we design coefficients to these time and money pieces.
-%		\item Describe why and how we design the two ratios: dollars per hour and dollars per aux.
-%	\end{itemize}
-%\end{itemize}
-
-A state is described as a set of \tit{state variables}.
-The state variables of a state $S$ include the transportation mode $M$ that led to $S$,
-time $T$ spent so far per mode (e.g., $T_\bike$ for biking and $T_\public$ for
-public transit), fare $D$ spent so far per mode (e.g., $D_\gas$ for driving and
-$D_\taxi$ for taking a cab), and variables related to the auxiliary data once uploaded.
-These extra data related variables are metrics such as the sum ($A_\SUM$),
-the maximum ($A_\MAX)$, the minimum ($A_\MIN$), and the average ($A_\AVG$) data along the path.
-
-We focused on weighted functions over state variables and
-designed the cost function, called preferential cost function (PCF), that guides the
-graph-based search engine in our trip planner as follows.
-\begin{equation}
-	\begin{aligned}
-		\PCF(S) =& \beta_1 * (\alpha_\walk \cdot T_\walk + \alpha_\bike \cdot T_\bike + \ldots) \\
-								&+ (D_\gas + D_\public + \ldots) \\
-								&+ \beta_2 * (A_\SUM + \ldots),
-	\end{aligned}
-	\label{eqt:pcf}
-\end{equation}
-where $\alpha_i$ are real numbers representing the relations among different time pieces,
-and $\beta_1$ ($\beta_2$) is the ratio that essentially describes how much in dollars a user would pay to
-save an hour (an auxiliary data, respectively).
-
-\subsubsection{Preference Elicitation}
-To gather these coefficients ($\alpha_i$'s and $\beta_i$'s) in our $\PCF$, we designed interface to
-elicit these numbers from the user.
-It asks the user questions and collect answers from the user to derive the coefficients.
-These questions are as follows.
-\begin{enumerate}
-	\item One hr of walking $=$ \underline{\hspace{1cm}} hrs of driving?
-	\item One hr of biking $=$ \underline{\hspace{1cm}} hrs of driving?
-	\item One hr of public transit $=$ \underline{\hspace{1cm}} hrs of driving?
-	\item One hr of taxi $=$ \underline{\hspace{1cm}} hrs of driving?
-	\item How much in dollars would you pay to save an hour in traveling?
-	\item How much in dollars would you pay to avoid an auxiliary event (e.g., crime) in traveling?
-\end{enumerate}
-For instance, Alice, an agent, answers 3, 2, 0.25 and 0.5 to the first four questions in the list above.
-Intuitively, the numbers indicate that she prefers public transit the most, followed by
-taxi, driving, biking and walking, in order.
-We show how we can now derive $\alpha_i$'s in \eqtref{pcf}.
-We start with setting $\alpha_\car=1$.
-Now, since one hour of walking is equivalent to 3 hours of walking,
-we have $\alpha_\walk \times 1 = \alpha_\car \times 3$;
-hence, we derive $\alpha_\walk = 3$.
-Similarly, we have $\alpha_\bike = 2$, $\alpha_\public = 0.25$,
-and $\alpha_\taxi = 0.5$.
-Computing $\beta_1$ and $\beta_1$ is straightforward.
-Indeed, function $\PCF$ with the input of time, fare and
-auxiliary metric pieces boils down to monetary cost,
-and the planner computes the best path by optimize
-based on this overall monetary cost in the searching
-process.
-
-\subsection{Reasoning with Constraints and Preferences}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Describe why and how we integrate constraints and preferences into A*.
-%\end{itemize}
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.4\textwidth]{figs/Astar.pdf}
-  \caption{Adjusted A*\label{fig:astar}}
-\end{figure}
-
-We leveraged the widely-used A* search algorithm on top of our high-performance graph search
-engine (cf. \figref{astar}).  The A* algorithm incorporates the following cost function.
-\begin{equation}
-	f(S) = g(S) + h(S),
-\end{equation}
-where $g(S)$ is the cost of an optimal trip from the initial state to $S$, and
-$h(S)$ is an admissible estimate of the cost of an optimal trip from $S$ to goal.
-
-To prune the search space, we check satisfiability of the temporal constraints in LTL
-at expansion of the search tree.
-To guide the search engine, we set $g(S)=\PCF(S)$ and $h(S)$ the minimum estimate among
-all available modes in $S$.
-
-%\section{User Study}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Describe the set up of our experiments on AMT and explain why: because we want to evaluate the
-%				feasibility of the three features.
-%	\item Describe the results.
-%\end{itemize}
-\section{Implementation and Results}
 %\begin{itemize}
@@
 
-\subsection{System Architecture}
+%\subsection{System Architecture}
 \begin{figure*}[!ht]
@@
 \end{figure}
-
-\subsection{Results}
-We present and analyze resulting routes for three agents:
-Alice, Bob, and Cal.
-This is assuming no auxiliary metric datasets uploaded
-so that the agents focus on time and fare.
-Fixing their where and when information, we show how
-agents' different constraints and preferences affect
-their optimal routes computed by our planner.
-Their where and when are set so that they all plan to
-travel from San Jose International Airport (SJC) in
-San Jose, to Pier 39 in San Francisco.
-Note that the natural constraint $\psi$ in \eqtref{ex} 
-is implicitly imposed on all cases.
-
-Agent Alice has only one constraint that she does not 
-have a car; thus, driving to her is never available during this trip.
-Per her preferences, Alice provides her thoughts as earlier;
-that is, public transportation is preferred to taxi, which is
-better than biking, preferred to walking.
-What's more, she decides that she would sacrifice thirty dollars
-to save one hour during the trip.
-
-The resulting route for Alice is included in \figref{alice}.
-The solution takes 2 hours 7 minutes and 18 dollars 94 cents.
-It is optimal in the sense the combined value of time
-and fare is minimal among all paths from SJC to Pier 39.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/result_Alice.pdf}
-  \caption{Resulting route for Alice\label{fig:alice}}
-\end{figure}
-
-Like Alice, Bob is constrained that he will not drive a car in his travel,
-and his dollar per hour is thirty.
-Moreover, Bob makes his mind to have some workout with his bike and
-dictates that he will bike for between one and two hours.
-He expresses his preferences: biking and public transit are the most
-preferred, next is taxi, and the least preferred is walking.
-Similarly, he has done so by answering the aforementioned elicitation questions.
-
-The result for Bob is depicted in \figref{bob}.
-It spans 2 hours 57 minutes in time with the fare of 29 dollars 17 cents.
-It is so, seemingly worse than what Alice achieved, only because Bob has
-the constraint that he will for sure bike for one to two hours, and
-the preferences that put biking the most satisfying mode.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/result_Bob.pdf}
-  \caption{Resulting route for Bob\label{fig:bob}}
-\end{figure}
-
-Agent Cal will also not use a car in the trip.
-He affirms that time is greatly valuable and one hour is measured
-to 500 dollars, although his budget is restricted to 50.
-Cal's preferences are that the most preferred are public transit
-and taxi, and the next preferred are walking and biking that
-are equivalent.
-
-Refer to \figref{cal} for the optimal path for Cal.
-This route, stretching 1 hour 48 minutes, is the most time-saving one 
-compared with the previous two, at the price of 49 dollars 91 cents.
-This is due to the constraint that Cal can spend up to 50 dollars,
-as well as his preferences and vale of time being high.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/result_Cal.pdf}
-  \caption{Resulting route for Cal\label{fig:cal}}
-\end{figure}
-
-\subsubsection{Auxiliary Metric}
-When the user of the planner is interested in metrics other than
-the ones offered already (i.e., time and fare), she might discover
-new metrics (e.g., crime rates and pollution statistics), upload
-them into the planner, and retrieve optimal plans taking these
-metrics into account.  One scenario of this approach is the following.
-
-Our agent needs to commute back home at night.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/crime.pdf}
-  \caption{Optimal route considering crime rates\label{fig:crime}}
-\end{figure}
-
-\section{Conclusion and Future Work}
-In multi-modal trip planning, it is vital for the planner to enable
-users to integrate new metrics of interest into the existing map,
-and to express constraints and preferences over which optimal
-routes are computed.
-In our work, we designed and built a graph-search based, extensible
-and personalizable multi-modal trip planner that utilizes linear
-temporal logic and preferential cost
-function to model user constraints and preferences.
-
-As a note on the future work, we plan to explore techniques in planning
-for computing multiple routes that are diverse with bounded difference of costs.
-Continuing this direction, we intend to study the problem of learning the
-$\PCF$ coefficients using the observations of the decisions the user made among
-the computed paths. (It is reasonable to assume that routes picked by the user
-are preferred to others, and these observations could contribute to the learning
-problem.)
-Also interesting is to introduce traffic information into the planner
-to support real-time multi-agent concurrent trip planning.
-
-\bibliographystyle{aaai}
-\bibliography{refs}
-
-\end{document}

### 2016-03-03 intro.tex
Broke paper.tex in pieces.

--- 2016-03-03 paper.tex
@@
-\def\year{2016}\relax
-%File: formatting-instruction.tex
-\documentclass[letterpaper]{article}
-\usepackage{aaai16}
-\usepackage{times}
-\usepackage{helvet}
-\usepackage{courier}
-
-\usepackage[pdftex]{hyperref}
-\usepackage{amsmath}
-\usepackage{color}
-\usepackage{textcomp}
-\usepackage{booktabs}
-\usepackage{ccicons}
-\usepackage{todonotes}
-\usepackage{amsfonts}
-\usepackage{framed}
-
-\frenchspacing
-\setlength{\pdfpagewidth}{8.5in}
-\setlength{\pdfpageheight}{11in}
-\pdfinfo{
-/Title (On Extensibility and Personalizability of Multi-Modal Trip Planning)
-/Author (Put All Your Authors Here, Separated by Commas)}
-\setcounter{secnumdepth}{0}
-
-\newcommand{\aux}{\mathit{aux}}
-\newcommand{\dist}{\mathit{dist}}
-\newcommand{\cA}{\mathcal{A}}
-\newcommand{\cU}{\mathcal{U}}
-\newcommand{\cP}{\mathcal{P}}
-\newcommand{\tit}[1]{\textit{#1}}
-\newcommand{\IFF}{\textit{iff}}
-\newcommand{\IF}{\textit{if}}
-\newcommand{\PCF}{\mathit{PCF}}
-\newcommand{\walk}{\mathit{walk}}
-\newcommand{\bike}{\mathit{bike}}
-\newcommand{\car}{\mathit{car}}
-\newcommand{\public}{\mathit{public}}
-\newcommand{\taxi}{\mathit{taxi}}
-\newcommand{\gas}{\mathit{gas}}
-\newcommand{\wait}{\mathit{wait}}
-\newcommand{\SUM}{\mathit{sum}}
-\newcommand{\MAX}{\mathit{max}}
-\newcommand{\MIN}{\mathit{min}}
-\newcommand{\AVG}{\mathit{avg}}
-
-\newcommand{\secref}[1]{Section~\ref{sec:#1}}
-\newcommand{\figref}[1]{Figure~\ref{fig:#1}}
-\newcommand{\tblref}[1]{Table~\ref{tbl:#1}}
-\newcommand{\algref}[1]{Algorithm~\ref{alg:#1}}
-\newcommand{\thmref}[1]{Theorem~\ref{thm:#1}}
-\newcommand{\propref}[1]{Proposition~\ref{prop:#1}}
-\newcommand{\corref}[1]{Corollary~\ref{cor:#1}}
-\newcommand{\eqtref}[1]{Equation~\ref{eqt:#1}}
-
- \begin{document}
-% The file aaai.sty is the style file for AAAI Press 
-% proceedings, working notes, and technical reports.
-%
-\title{On Extensibility and Personalizability of Multi-Modal Trip Planning}
-%\author{
-%	Xudong Liu\\
-%	Department of Computer Science\\
-%	University of Kentucky\\
-%	329 Rose Street\\
-%	Lexington, KY 40506\\
-%	USA
-%	\And
-%	PARC People\\
-%	Palo Alto Research Center\\
-%	3333 Coyote Hill Road\\
-%	Palo Alto, CA 94304\\
-%	USA
-%}
-\maketitle
-\begin{abstract}
-	Trip planning is a practical task that has drawn
-	extensive attention from the AI planning and scheduling community
-	and the industrial/commercial sectors.
-	In this paper, we consider the setting of the \tit{multi-modal}
-	trip planning, where users can exploit different transportation
-	modes, such as walking, biking, public transit, and taxi.
-	In such a context, it would be of benefit if the
-	user was able to extend the \tit{cost base}, including traveling time 
-	and fare, of the planner, and to personalize the planner according
-	to her own constraints and preferences.
-	To this end, we designed and developed a hipergraph-based multi-modal
-	trip planner that allows users to upload auxiliary cost metrics (e.g.,
-	crime rates), to specify constraints as a theory in the 
-	\tit{linear temporal logic}, and to express preferences as a 
-	\tit{preferential cost function}.
-\end{abstract}
-
-\section{Introduction}
 Trip planning, as an application of planning and scheduling,
@@
 %\end{itemize}
-
-\section{Extensibility}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Explain in detail what extensibility means.
-%	\item Describe the input.
-%	\item Describe how we incorporate the auxiliary data into the planner.
-%\end{itemize}
-Allowing users to upload their own data sets of interests is an important
-step towards customization of a trip planner.
-We designed a framework where a user can upload auxiliary cost 
-metric data (e.g., crime
-statistics and pollution data) into the planner,
-and the planner will compute an optimal route accordingly.
-
-The user-created data are the auxiliary data that is represented as pairs
-of latitude and longitude degrees.
-To merge these lat-long pairs into the planner, we performed a neighborhood
-search to calculate the total score of auxiliary data for each 
-lat-long pair already in our planner.
-It might be of strong interest to some user for our planning
-system to take care of criminal statistics so that some level of safety
-of the resulting routes is guaranteed.
-For instance, a user traveling through the downtown area of San Francisco
-around midnight may want to upload a data set of crimes (cf. \figref{sf_crime}), 
-and express her constraints and preferences in hope of a preferred trip plan.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.4\textwidth]{figs/sub_sf_crime_July2015.png}
-  \caption{Crime rates in San Francisco\label{fig:sf_crime}}
-\end{figure}
-
-Formally, let $\cU$ be the uploaded auxiliary dataset of lat-long pairs, 
-$N=(x_N,y_N)$ a node in our planner and $r$ the effective radius.
-We first compute the set $\cP \subseteq A$ of auxiliary points such that,
-for every node $P \in \cP$, the Euclidean distance between $N$ and $P$
-is within $r$.
-Then, the score of auxiliary data for $N$ with respect to $\cP$ and $r$,
-denoted by $S(N,\cP,r)$, is computed as follows.
-
-\begin{equation*}
-	S(N,\cP,r) = \sum_{\substack{P \in \cP}} 
-		\Big(1 - \frac{\sqrt{(x_N-x_P)^2+(y_N-y_P)^2}}{r}\Big)
-\end{equation*}
-
-Now we turn to \figref{aux} for an instance to show how auxiliary data
-are integrated into the graph using the equation above.
-In \figref{aux}, we have the green nodes denoting the graph nodes, 
-and the red nodes the new auxiliary nodes that represent locations
-of criminal events.
-For node $N$, there are three red nodes within its neighborhood of
-radius of $r$; thus, we have $\cP=\{P_1,P_2,P_3\}$.
-We assume $r$ is 100 feet, and the distances from $N$ to $P_1$,
-$P_2$ and $P_3$ are 25, 80 and 90 feet, respectively.
-Then, the crime score of $S(N,\cP,r)$ is $1.05$.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.4\textwidth]{figs/aux.pdf}
-  \caption{Integrating uploaded data\label{fig:aux}}
-\end{figure}
-
-\section{Personalizability}
-Personalizability consists of two aspects: constraints and preferences.
-From the viewpoint of the planner,
-constraints, also referred to as hard constraints, are statements that the planner
-has to satisfy during the planning process; whereas preferences, also called
-soft constraints, are specifications that the planner will need to optimize.
-We formulated constraints using linear temporal logic (LTL) and preferences as
-a preferential cost function (PCF), and implemented our planner leveraging the
-widely-used graph search algorithm the A*.
-
-\subsection{Constraints}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Describe syntax and semantics of LTL in the setting of trip planning.
-%	\begin{itemize}
-%		\setlength\itemsep{0pt}
-%		\item Specifically, we need to point out the ``after" temporal connective,
-%					and we need to describe the semantics: what is a plan, why we can
-%					eliminate actions and put them as state attributes instead, etc.
-%	\end{itemize}
-%\end{itemize}
-As constraints in the setting of trip planning are often declarative and
-temporal, our choice of LTL is straightforward.
-We now give a brief review of linear temporal logic (LTL).
-Let $f$ be a propositional formula over a finite set $L$ of Boolean variables.  
-LTL formulas are defined recursively as follows.
-\begin{equation}
-	\varphi = f | \varphi_1 \land \varphi_2 | \varphi_1 \lor \varphi_2 | \neg \varphi | 
-		\bigcirc \varphi |	\Box \varphi | \Diamond \varphi | \varphi_1 \cA \varphi_2
-\end{equation}
-Note that we have $\varphi_1 \cA \varphi_2$, and it means that
-``$\varphi_2$ holds right after $\varphi_1$ holds."
-
-A natural constraint in trip planning could be ``In this trip I will not drive a car 
-after biking or taking the public transit."
-In LTL, such constraint can be translated into an LTL formula
-\begin{equation}
-\label{eqt:ex}
-	\psi = ((M=b) \lor (M=p)) \,\cA\, (\Box (\neg (M=c))).
-\end{equation}
-
-As the actions in trip planning is limited to taking different transportation modes,
-in our definition of the semantics of LTL
-these actions are subsumed into the interpretations of $L$, or \tit{states}.
-The semantics of LTL is defined with regard to trajectories of states. 
-Let $\sigma$ be a trajectory of states $S_0,a_1,S_1,\ldots,a_n,S_n$, and
-$\sigma[i]$ a suffix $S_i, a_{i+1}, S_{i+1}, \ldots,a_n,S_n$.  We have
-\begin{align*}
-	\sigma \models f \;\; &\IFF \;\; S_0 \models f,\\
-	\sigma \models \varphi_1 \land \varphi_2 \;\; &\IFF \;\; \sigma \models \varphi_1 \; and \; \sigma \models \varphi_2,\\
-	\sigma \models \varphi_1 \lor \varphi_2 \;\; &\IFF \;\; \sigma \models \varphi_1 \; or \; \sigma \models \varphi_2,\\
-	\sigma \models \neg \varphi \;\; &\IFF \;\; \sigma \not \models \varphi,\\
-	\sigma \models \bigcirc \varphi \;\; &\IFF \;\; \sigma[1] \models \varphi,\\
-	\sigma \models \Box \varphi \;\; &\IFF \;\; \forall 0 \leq i \leq n (\sigma[i] \models \varphi),\\
-	\sigma \models \Diamond \varphi \;\; &\IFF \;\; \exists 0 \leq i \leq n (\sigma[i] \models \varphi),\\
-	\sigma \models \varphi_1 \cA \varphi_2 \;\; &\IFF \;\; \forall 0 \leq i < n (\IF \; \sigma[i] \models \varphi_1, \sigma[i+1] \models \varphi_2).
-\end{align*}
-
-For example, we are given an LTL constraint $\psi$ and three trajectories ($\sigma_1$, $\sigma_2$ and $\sigma_3$)
-as shown in \figref{trjs}.
-Clearly, we have $\sigma_1 \models \psi$ because, after public transit in $S_2$ and $S_3$, 
-traveling by car has never been taken place.
-Moreover, we have $\sigma_2 \not \models \psi$ because we have $M=c$ hold in $S_6$ and $S_7$
-after having $M=p$ hold in $S_2$ and $S_5$.
-Finally, we know $\sigma_3 \models \psi$, as the mode is always neither biking nor public transit.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.4\textwidth]{figs/trajectories.pdf}
-  \caption{State transition diagram\label{fig:trjs}}
-\end{figure}
-
-\subsection{Preferences}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Describe our preferential cost function (PCF).
-%	\begin{itemize}
-%		\setlength\itemsep{0pt}
-%		\item Describe why and how we break down time and money.
-%		\item Describe why and how we design coefficients to these time and money pieces.
-%		\item Describe why and how we design the two ratios: dollars per hour and dollars per aux.
-%	\end{itemize}
-%\end{itemize}
-
-A state is described as a set of \tit{state variables}.
-The state variables of a state $S$ include the transportation mode $M$ that led to $S$,
-time $T$ spent so far per mode (e.g., $T_\bike$ for biking and $T_\public$ for
-public transit), fare $D$ spent so far per mode (e.g., $D_\gas$ for driving and
-$D_\taxi$ for taking a cab), and variables related to the auxiliary data once uploaded.
-These extra data related variables are metrics such as the sum ($A_\SUM$),
-the maximum ($A_\MAX)$, the minimum ($A_\MIN$), and the average ($A_\AVG$) data along the path.
-
-We focused on weighted functions over state variables and
-designed the cost function, called preferential cost function (PCF), that guides the
-graph-based search engine in our trip planner as follows.
-\begin{equation}
-	\begin{aligned}
-		\PCF(S) =& \beta_1 * (\alpha_\walk \cdot T_\walk + \alpha_\bike \cdot T_\bike + \ldots) \\
-								&+ (D_\gas + D_\public + \ldots) \\
-								&+ \beta_2 * (A_\SUM + \ldots),
-	\end{aligned}
-	\label{eqt:pcf}
-\end{equation}
-where $\alpha_i$ are real numbers representing the relations among different time pieces,
-and $\beta_1$ ($\beta_2$) is the ratio that essentially describes how much in dollars a user would pay to
-save an hour (an auxiliary data, respectively).
-
-\subsubsection{Preference Elicitation}
-To gather these coefficients ($\alpha_i$'s and $\beta_i$'s) in our $\PCF$, we designed interface to
-elicit these numbers from the user.
-It asks the user questions and collect answers from the user to derive the coefficients.
-These questions are as follows.
-\begin{enumerate}
-	\item One hr of walking $=$ \underline{\hspace{1cm}} hrs of driving?
-	\item One hr of biking $=$ \underline{\hspace{1cm}} hrs of driving?
-	\item One hr of public transit $=$ \underline{\hspace{1cm}} hrs of driving?
-	\item One hr of taxi $=$ \underline{\hspace{1cm}} hrs of driving?
-	\item How much in dollars would you pay to save an hour in traveling?
-	\item How much in dollars would you pay to avoid an auxiliary event (e.g., crime) in traveling?
-\end{enumerate}
-For instance, Alice, an agent, answers 3, 2, 0.25 and 0.5 to the first four questions in the list above.
-Intuitively, the numbers indicate that she prefers public transit the most, followed by
-taxi, driving, biking and walking, in order.
-We show how we can now derive $\alpha_i$'s in \eqtref{pcf}.
-We start with setting $\alpha_\car=1$.
-Now, since one hour of walking is equivalent to 3 hours of walking,
-we have $\alpha_\walk \times 1 = \alpha_\car \times 3$;
-hence, we derive $\alpha_\walk = 3$.
-Similarly, we have $\alpha_\bike = 2$, $\alpha_\public = 0.25$,
-and $\alpha_\taxi = 0.5$.
-Computing $\beta_1$ and $\beta_1$ is straightforward.
-Indeed, function $\PCF$ with the input of time, fare and
-auxiliary metric pieces boils down to monetary cost,
-and the planner computes the best path by optimize
-based on this overall monetary cost in the searching
-process.
-
-\subsection{Reasoning with Constraints and Preferences}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Describe why and how we integrate constraints and preferences into A*.
-%\end{itemize}
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.4\textwidth]{figs/Astar.pdf}
-  \caption{Adjusted A*\label{fig:astar}}
-\end{figure}
-
-We leveraged the widely-used A* search algorithm on top of our high-performance graph search
-engine (cf. \figref{astar}).  The A* algorithm incorporates the following cost function.
-\begin{equation}
-	f(S) = g(S) + h(S),
-\end{equation}
-where $g(S)$ is the cost of an optimal trip from the initial state to $S$, and
-$h(S)$ is an admissible estimate of the cost of an optimal trip from $S$ to goal.
-
-To prune the search space, we check satisfiability of the temporal constraints in LTL
-at expansion of the search tree.
-To guide the search engine, we set $g(S)=\PCF(S)$ and $h(S)$ the minimum estimate among
-all available modes in $S$.
-
-%\section{User Study}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Describe the set up of our experiments on AMT and explain why: because we want to evaluate the
-%				feasibility of the three features.
-%	\item Describe the results.
-%\end{itemize}
-\section{Implementation and Results}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Describe the system structure.
-%	\item Describe the input to the system.
-%	\item Describe the output from the system.
-%\end{itemize}
-
-\subsection{System Architecture}
-\begin{figure*}[!ht]
-  \centering
-    \includegraphics[width=0.8\textwidth]{figs/system.pdf}
-  \caption{System Overview\label{fig:system}}
-\end{figure*}
-We designed and implemented a multi-modal trip planning
-system based on a high-performance graph search engine.
-The planner allows user uploads, as well as declarative
-constraints and preferences.
-We now describe the structure of the planning system.
-
-The trip planner takes two types of data as input:
-static data and user-specified request.
-The static input includes Map Data and Transit
-Data.
-Map Data describes the map, a directed graph where 
-nodes are street corners, bus stops and train stations.
-Transit Data is a set of schedules for the buses and trains
-On the other hand, a user provides her request, composed of 
-three parts.
-First, the user enters \tit{from} and \tit{to} locations 
-on the map together with day and time of the start of the trip.  
-Second, the user may upload her auxiliary metric dataset, e.g., crime rates.  
-Lastly, the user specifies her constraints in LTL and preferences as a $\PCF$.
-For example, the constraint 
-could be ``never walk through a bad neighborhood."  Given these 
-inputs, our planner computes an optimal path satisfying 
-all the constraints and optimizing the preferences.
-Note that the request from the user is encapsulated into
-a JSON object.
-
-For instance, the JSON object for the constraint $\psi$ in \eqtref{ex}
-is shown in \figref{exJSON}.
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/exampleJSON.png}
-  \caption{JSON object for constraint $\psi$ in \eqtref{ex}\label{fig:exJSON}}
-\end{figure}
-
-\subsection{Results}
-We present and analyze resulting routes for three agents:
-Alice, Bob, and Cal.
-This is assuming no auxiliary metric datasets uploaded
-so that the agents focus on time and fare.
-Fixing their where and when information, we show how
-agents' different constraints and preferences affect
-their optimal routes computed by our planner.
-Their where and when are set so that they all plan to
-travel from San Jose International Airport (SJC) in
-San Jose, to Pier 39 in San Francisco.
-Note that the natural constraint $\psi$ in \eqtref{ex} 
-is implicitly imposed on all cases.
-
-Agent Alice has only one constraint that she does not 
-have a car; thus, driving to her is never available during this trip.
-Per her preferences, Alice provides her thoughts as earlier;
-that is, public transportation is preferred to taxi, which is
-better than biking, preferred to walking.
-What's more, she decides that she would sacrifice thirty dollars
-to save one hour during the trip.
-
-The resulting route for Alice is included in \figref{alice}.
-The solution takes 2 hours 7 minutes and 18 dollars 94 cents.
-It is optimal in the sense the combined value of time
-and fare is minimal among all paths from SJC to Pier 39.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/result_Alice.pdf}
-  \caption{Resulting route for Alice\label{fig:alice}}
-\end{figure}
-
-Like Alice, Bob is constrained that he will not drive a car in his travel,
-and his dollar per hour is thirty.
-Moreover, Bob makes his mind to have some workout with his bike and
-dictates that he will bike for between one and two hours.
-He expresses his preferences: biking and public transit are the most
-preferred, next is taxi, and the least preferred is walking.
-Similarly, he has done so by answering the aforementioned elicitation questions.
-
-The result for Bob is depicted in \figref{bob}.
-It spans 2 hours 57 minutes in time with the fare of 29 dollars 17 cents.
-It is so, seemingly worse than what Alice achieved, only because Bob has
-the constraint that he will for sure bike for one to two hours, and
-the preferences that put biking the most satisfying mode.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/result_Bob.pdf}
-  \caption{Resulting route for Bob\label{fig:bob}}
-\end{figure}
-
-Agent Cal will also not use a car in the trip.
-He affirms that time is greatly valuable and one hour is measured
-to 500 dollars, although his budget is restricted to 50.
-Cal's preferences are that the most preferred are public transit
-and taxi, and the next preferred are walking and biking that
-are equivalent.
-
-Refer to \figref{cal} for the optimal path for Cal.
-This route, stretching 1 hour 48 minutes, is the most time-saving one 
-compared with the previous two, at the price of 49 dollars 91 cents.
-This is due to the constraint that Cal can spend up to 50 dollars,
-as well as his preferences and vale of time being high.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/result_Cal.pdf}
-  \caption{Resulting route for Cal\label{fig:cal}}
-\end{figure}
-
-\subsubsection{Auxiliary Metric}
-When the user of the planner is interested in metrics other than
-the ones offered already (i.e., time and fare), she might discover
-new metrics (e.g., crime rates and pollution statistics), upload
-them into the planner, and retrieve optimal plans taking these
-metrics into account.  One scenario of this approach is the following.
-
-Our agent needs to commute back home at night.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/crime.pdf}
-  \caption{Optimal route considering crime rates\label{fig:crime}}
-\end{figure}
-
-\section{Conclusion and Future Work}
-In multi-modal trip planning, it is vital for the planner to enable
-users to integrate new metrics of interest into the existing map,
-and to express constraints and preferences over which optimal
-routes are computed.
-In our work, we designed and built a graph-search based, extensible
-and personalizable multi-modal trip planner that utilizes linear
-temporal logic and preferential cost
-function to model user constraints and preferences.
-
-As a note on the future work, we plan to explore techniques in planning
-for computing multiple routes that are diverse with bounded difference of costs.
-Continuing this direction, we intend to study the problem of learning the
-$\PCF$ coefficients using the observations of the decisions the user made among
-the computed paths. (It is reasonable to assume that routes picked by the user
-are preferred to others, and these observations could contribute to the learning
-problem.)
-Also interesting is to introduce traffic information into the planner
-to support real-time multi-agent concurrent trip planning.
-
-\bibliographystyle{aaai}
-\bibliography{refs}
-
-\end{document}

### 2016-03-03 paper.tex
Broke paper.tex in pieces.

--- 2016-03-03 paper.tex
@@
 \section{Introduction}
-Trip planning, as an application of planning and scheduling,
-has seen substantive implementations by researchers and
-developers\cite{bast2015route}.
-Some of the planning systems are \tit{multi-modal}; that is,
-combining distinct transportation modes, the trip planners
-compute optimal routes from sources to destinations.
-This notion of ``optimality" generally refers to the computed routes
-having minimal total time or total fare.
-
-However, in the eyes of a user, it may be more faceted than just
-``fastest" or ``cheapest."
-For instance, for a college student who specifies that she will
-only walk or take public transit in a trip from Palo Alto to
-San Francisco, the computed plan is not necessarily the fastest
-(e.g., taking a cab could be faster)
-or the cheapest (e.g., walking all the way has no fare).
-This happens when a user tells the planner her hard constraints,
-called \tit{constraints}.  
-The planner then needs to either satisfy the 
-constraints in the search process or return failure because they are
-over-restrictive.
-Moreover, the user might want to further customize the planner by
-describing soft constraints, called \tit{preferences}.
-For example, an agent wants to travel from school to
-downtown, and prefers biking to taking the bus.
-Thus, a trip with more biking than bus may be considered better for
-the agent than the one with more bus than biking.
-In this case, the planner will need to accommodate user preferences
-whenever possible in the search of optimal solutions.
-
-Representing and reasoning about constraints and preferences are
-fundamental to decision making in automated planning and scheduling
-in artificial intelligence.
-Son and Pontelli presented a declarative preference language, 
-$\mathcal{PP}$, to represent ordinal preferences (e.g., basic desires,
-atomic preferences and general preferences) between trajectories of 
-a planning problem\cite{son2004planning}.
-Bienvenu et al. proposed a first-order preference language, $\mathcal{LPP}$, 
-extending $\mathcal{PP}$ by allowing the specification of quantified qualitative
-preferences\cite{bienvenu2011specifying}.
-Modesti and Sciomachen introduced an ad hoc utility function for weighing
-the arcs both with their time and fare\cite{modesti1998utility}.
-This relationship between time and fare in the setting of transportation
-was studied, and economic models capturing the trade-off between the two
-was presented\cite{antoniou2007methodology}.
-
-However, relatively limited effort has been devoted to designing and
-implementing real-world multi-modal trip planners that captures user
-constraints and preferences over the \tit{cost base}, possibly
-extended from the user with auxiliary \tit{cost metrics},
-such as crime rates and pollution statistics.
-
-Using a high-performance graph search
-engine\cite{zhou2011dynamic}, we designed and implemented a 
-multi-modal trip planner that uses pure
-graph-search. This allows us to flexibly combine various modes
-(e.g., walk, bike, car, public transit, and taxi) and 
-declaratively change constraints and preferences. 
-The planner allows the user
-to upload \tit{new} mapping data over which to express
-constraints and preferences.  For instance, a user might upload a map of crime
-in the city, and ask the trip planner to avoid areas where crime is frequent.
-Then, the planner takes constraints expressed in \tit{linear
-temporal logic} to constrain the search space (e.g., never bike after
-transit, and never walk through bad neighborhoods).
-Finally, the planner uses a weighted sum, called \tit{preferential cost function}, 
-over several cost metrics (e.g., time spend biking, fare on public transit, 
-and overall crimes walking through) which can be re-weighted based
-on different user preferences.
-
-Our paper is organized as follows.
-In the next section, we present what it means for a planner to be extensible
-and formally define the method to incorporating new metrics into the planner.
-In the following section, we discuss the two aspects of personalization
-in trip planning: constraints and preferences, and how they are modeled and
-reasoned with in the setting of multi-modal trip planning.
-We then move on to describe the system structure of our graph-search based planner,
-and show results obtained from our planner in various occasions.
-Finally, we conclude by outlining some future research directions.
-
-%\begin{itemize}
-%	\item What's the problem: personalized trip planning: find trips that
-%	  are optimal in the eyes of a user, who doesn't care about mode or
-%	  provider boundaries, and whose preferences may be more faceted than
-%	  ``fastest'' or ``cheapest''.
-%	\item Why is it hard: the computational complexity of searching
-%	  combinations of multiple modes that can be combined freely is high
-%	  (PSPACE-complete\cite{}) which is why existing trip planner exploit problem
-%	  structure (such as only one mode per trip) and predefined
-%	  preferences (optimization metrics)
-%	\item How did we solve it: Using a high-performance graph search
-%	  engine\cite{}, we implemented a multi-modal trip planner that uses pure
-%	  graph-search. This allows us to flexibly combine various modes
-%	  (e.g., car, train, cab, walk) and declarative change soft and hard
-%	  constraints. The planner can take constraints expressed in linear
-%	  temporal logic to constrain the search space (e.g., never bike after
-%	  transit, i.e., I don't want to take my bike on the bus or train).
-%	  The planner uses a weighted sum over several metrics (e.g., time
-%	  overall, time spend on transit, cost) which can be re-weighted based
-%	  on different user preferences. Finally the planner allows the user
-%	  to upload {\em new} mapping data over which to express hard and/or
-%	  soft constraints. For instance, a user might upload a map of crime
-%	  in the city, and ask the trip planner to never walk at night through
-%	  an area where crime is frequent.
-%%	\item How did we evaluate the features and what are the results:
-%%	  ceteris paribus experimentation setup on Amazon Mechanical 
-%%		Turk (MTurk\footnote{\url{http://mturk.amazon.com}}) and results, 
-%%		asking users to pick between two different trips, one being limited by the
-%%	  capabilities of existing planners, and one exhibiting features only
-%%	  available in our trip planner.
-%\end{itemize}
+\input{intro}
 
 \section{Extensibility}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Explain in detail what extensibility means.
-%	\item Describe the input.
-%	\item Describe how we incorporate the auxiliary data into the planner.
-%\end{itemize}
-Allowing users to upload their own data sets of interests is an important
-step towards customization of a trip planner.
-We designed a framework where a user can upload auxiliary cost 
-metric data (e.g., crime
-statistics and pollution data) into the planner,
-and the planner will compute an optimal route accordingly.
-
-The user-created data are the auxiliary data that is represented as pairs
-of latitude and longitude degrees.
-To merge these lat-long pairs into the planner, we performed a neighborhood
-search to calculate the total score of auxiliary data for each 
-lat-long pair already in our planner.
-It might be of strong interest to some user for our planning
-system to take care of criminal statistics so that some level of safety
-of the resulting routes is guaranteed.
-For instance, a user traveling through the downtown area of San Francisco
-around midnight may want to upload a data set of crimes (cf. \figref{sf_crime}), 
-and express her constraints and preferences in hope of a preferred trip plan.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.4\textwidth]{figs/sub_sf_crime_July2015.png}
-  \caption{Crime rates in San Francisco\label{fig:sf_crime}}
-\end{figure}
-
-Formally, let $\cU$ be the uploaded auxiliary dataset of lat-long pairs, 
-$N=(x_N,y_N)$ a node in our planner and $r$ the effective radius.
-We first compute the set $\cP \subseteq A$ of auxiliary points such that,
-for every node $P \in \cP$, the Euclidean distance between $N$ and $P$
-is within $r$.
-Then, the score of auxiliary data for $N$ with respect to $\cP$ and $r$,
-denoted by $S(N,\cP,r)$, is computed as follows.
-
-\begin{equation*}
-	S(N,\cP,r) = \sum_{\substack{P \in \cP}} 
-		\Big(1 - \frac{\sqrt{(x_N-x_P)^2+(y_N-y_P)^2}}{r}\Big)
-\end{equation*}
-
-Now we turn to \figref{aux} for an instance to show how auxiliary data
-are integrated into the graph using the equation above.
-In \figref{aux}, we have the green nodes denoting the graph nodes, 
-and the red nodes the new auxiliary nodes that represent locations
-of criminal events.
-For node $N$, there are three red nodes within its neighborhood of
-radius of $r$; thus, we have $\cP=\{P_1,P_2,P_3\}$.
-We assume $r$ is 100 feet, and the distances from $N$ to $P_1$,
-$P_2$ and $P_3$ are 25, 80 and 90 feet, respectively.
-Then, the crime score of $S(N,\cP,r)$ is $1.05$.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.4\textwidth]{figs/aux.pdf}
-  \caption{Integrating uploaded data\label{fig:aux}}
-\end{figure}
+\input{extensibility}
 
 \section{Personalizability}
-Personalizability consists of two aspects: constraints and preferences.
-From the viewpoint of the planner,
-constraints, also referred to as hard constraints, are statements that the planner
-has to satisfy during the planning process; whereas preferences, also called
-soft constraints, are specifications that the planner will need to optimize.
-We formulated constraints using linear temporal logic (LTL) and preferences as
-a preferential cost function (PCF), and implemented our planner leveraging the
-widely-used graph search algorithm the A*.
-
-\subsection{Constraints}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Describe syntax and semantics of LTL in the setting of trip planning.
-%	\begin{itemize}
-%		\setlength\itemsep{0pt}
-%		\item Specifically, we need to point out the ``after" temporal connective,
-%					and we need to describe the semantics: what is a plan, why we can
-%					eliminate actions and put them as state attributes instead, etc.
-%	\end{itemize}
-%\end{itemize}
-As constraints in the setting of trip planning are often declarative and
-temporal, our choice of LTL is straightforward.
-We now give a brief review of linear temporal logic (LTL).
-Let $f$ be a propositional formula over a finite set $L$ of Boolean variables.  
-LTL formulas are defined recursively as follows.
-\begin{equation}
-	\varphi = f | \varphi_1 \land \varphi_2 | \varphi_1 \lor \varphi_2 | \neg \varphi | 
-		\bigcirc \varphi |	\Box \varphi | \Diamond \varphi | \varphi_1 \cA \varphi_2
-\end{equation}
-Note that we have $\varphi_1 \cA \varphi_2$, and it means that
-``$\varphi_2$ holds right after $\varphi_1$ holds."
-
-A natural constraint in trip planning could be ``In this trip I will not drive a car 
-after biking or taking the public transit."
-In LTL, such constraint can be translated into an LTL formula
-\begin{equation}
-\label{eqt:ex}
-	\psi = ((M=b) \lor (M=p)) \,\cA\, (\Box (\neg (M=c))).
-\end{equation}
-
-As the actions in trip planning is limited to taking different transportation modes,
-in our definition of the semantics of LTL
-these actions are subsumed into the interpretations of $L$, or \tit{states}.
-The semantics of LTL is defined with regard to trajectories of states. 
-Let $\sigma$ be a trajectory of states $S_0,a_1,S_1,\ldots,a_n,S_n$, and
-$\sigma[i]$ a suffix $S_i, a_{i+1}, S_{i+1}, \ldots,a_n,S_n$.  We have
-\begin{align*}
-	\sigma \models f \;\; &\IFF \;\; S_0 \models f,\\
-	\sigma \models \varphi_1 \land \varphi_2 \;\; &\IFF \;\; \sigma \models \varphi_1 \; and \; \sigma \models \varphi_2,\\
-	\sigma \models \varphi_1 \lor \varphi_2 \;\; &\IFF \;\; \sigma \models \varphi_1 \; or \; \sigma \models \varphi_2,\\
-	\sigma \models \neg \varphi \;\; &\IFF \;\; \sigma \not \models \varphi,\\
-	\sigma \models \bigcirc \varphi \;\; &\IFF \;\; \sigma[1] \models \varphi,\\
-	\sigma \models \Box \varphi \;\; &\IFF \;\; \forall 0 \leq i \leq n (\sigma[i] \models \varphi),\\
-	\sigma \models \Diamond \varphi \;\; &\IFF \;\; \exists 0 \leq i \leq n (\sigma[i] \models \varphi),\\
-	\sigma \models \varphi_1 \cA \varphi_2 \;\; &\IFF \;\; \forall 0 \leq i < n (\IF \; \sigma[i] \models \varphi_1, \sigma[i+1] \models \varphi_2).
-\end{align*}
-
-For example, we are given an LTL constraint $\psi$ and three trajectories ($\sigma_1$, $\sigma_2$ and $\sigma_3$)
-as shown in \figref{trjs}.
-Clearly, we have $\sigma_1 \models \psi$ because, after public transit in $S_2$ and $S_3$, 
-traveling by car has never been taken place.
-Moreover, we have $\sigma_2 \not \models \psi$ because we have $M=c$ hold in $S_6$ and $S_7$
-after having $M=p$ hold in $S_2$ and $S_5$.
-Finally, we know $\sigma_3 \models \psi$, as the mode is always neither biking nor public transit.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.4\textwidth]{figs/trajectories.pdf}
-  \caption{State transition diagram\label{fig:trjs}}
-\end{figure}
-
-\subsection{Preferences}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Describe our preferential cost function (PCF).
-%	\begin{itemize}
-%		\setlength\itemsep{0pt}
-%		\item Describe why and how we break down time and money.
-%		\item Describe why and how we design coefficients to these time and money pieces.
-%		\item Describe why and how we design the two ratios: dollars per hour and dollars per aux.
-%	\end{itemize}
-%\end{itemize}
-
-A state is described as a set of \tit{state variables}.
-The state variables of a state $S$ include the transportation mode $M$ that led to $S$,
-time $T$ spent so far per mode (e.g., $T_\bike$ for biking and $T_\public$ for
-public transit), fare $D$ spent so far per mode (e.g., $D_\gas$ for driving and
-$D_\taxi$ for taking a cab), and variables related to the auxiliary data once uploaded.
-These extra data related variables are metrics such as the sum ($A_\SUM$),
-the maximum ($A_\MAX)$, the minimum ($A_\MIN$), and the average ($A_\AVG$) data along the path.
-
-We focused on weighted functions over state variables and
-designed the cost function, called preferential cost function (PCF), that guides the
-graph-based search engine in our trip planner as follows.
-\begin{equation}
-	\begin{aligned}
-		\PCF(S) =& \beta_1 * (\alpha_\walk \cdot T_\walk + \alpha_\bike \cdot T_\bike + \ldots) \\
-								&+ (D_\gas + D_\public + \ldots) \\
-								&+ \beta_2 * (A_\SUM + \ldots),
-	\end{aligned}
-	\label{eqt:pcf}
-\end{equation}
-where $\alpha_i$ are real numbers representing the relations among different time pieces,
-and $\beta_1$ ($\beta_2$) is the ratio that essentially describes how much in dollars a user would pay to
-save an hour (an auxiliary data, respectively).
-
-\subsubsection{Preference Elicitation}
-To gather these coefficients ($\alpha_i$'s and $\beta_i$'s) in our $\PCF$, we designed interface to
-elicit these numbers from the user.
-It asks the user questions and collect answers from the user to derive the coefficients.
-These questions are as follows.
-\begin{enumerate}
-	\item One hr of walking $=$ \underline{\hspace{1cm}} hrs of driving?
-	\item One hr of biking $=$ \underline{\hspace{1cm}} hrs of driving?
-	\item One hr of public transit $=$ \underline{\hspace{1cm}} hrs of driving?
-	\item One hr of taxi $=$ \underline{\hspace{1cm}} hrs of driving?
-	\item How much in dollars would you pay to save an hour in traveling?
-	\item How much in dollars would you pay to avoid an auxiliary event (e.g., crime) in traveling?
-\end{enumerate}
-For instance, Alice, an agent, answers 3, 2, 0.25 and 0.5 to the first four questions in the list above.
-Intuitively, the numbers indicate that she prefers public transit the most, followed by
-taxi, driving, biking and walking, in order.
-We show how we can now derive $\alpha_i$'s in \eqtref{pcf}.
-We start with setting $\alpha_\car=1$.
-Now, since one hour of walking is equivalent to 3 hours of walking,
-we have $\alpha_\walk \times 1 = \alpha_\car \times 3$;
-hence, we derive $\alpha_\walk = 3$.
-Similarly, we have $\alpha_\bike = 2$, $\alpha_\public = 0.25$,
-and $\alpha_\taxi = 0.5$.
-Computing $\beta_1$ and $\beta_1$ is straightforward.
-Indeed, function $\PCF$ with the input of time, fare and
-auxiliary metric pieces boils down to monetary cost,
-and the planner computes the best path by optimize
-based on this overall monetary cost in the searching
-process.
-
-\subsection{Reasoning with Constraints and Preferences}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Describe why and how we integrate constraints and preferences into A*.
-%\end{itemize}
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.4\textwidth]{figs/Astar.pdf}
-  \caption{Adjusted A*\label{fig:astar}}
-\end{figure}
-
-We leveraged the widely-used A* search algorithm on top of our high-performance graph search
-engine (cf. \figref{astar}).  The A* algorithm incorporates the following cost function.
-\begin{equation}
-	f(S) = g(S) + h(S),
-\end{equation}
-where $g(S)$ is the cost of an optimal trip from the initial state to $S$, and
-$h(S)$ is an admissible estimate of the cost of an optimal trip from $S$ to goal.
-
-To prune the search space, we check satisfiability of the temporal constraints in LTL
-at expansion of the search tree.
-To guide the search engine, we set $g(S)=\PCF(S)$ and $h(S)$ the minimum estimate among
-all available modes in $S$.
+\input{personalizability}
 
@@
 %\end{itemize}
-\section{Implementation and Results}
-%\begin{itemize}
-%	\setlength\itemsep{1pt}
-%	\item Describe the system structure.
-%	\item Describe the input to the system.
-%	\item Describe the output from the system.
-%\end{itemize}
 
-\subsection{System Architecture}
-\begin{figure*}[!ht]
-  \centering
-    \includegraphics[width=0.8\textwidth]{figs/system.pdf}
-  \caption{System Overview\label{fig:system}}
-\end{figure*}
-We designed and implemented a multi-modal trip planning
-system based on a high-performance graph search engine.
-The planner allows user uploads, as well as declarative
-constraints and preferences.
-We now describe the structure of the planning system.
+\section{Implementation}
+\input{implementation}
 
-The trip planner takes two types of data as input:
-static data and user-specified request.
-The static input includes Map Data and Transit
-Data.
-Map Data describes the map, a directed graph where 
-nodes are street corners, bus stops and train stations.
-Transit Data is a set of schedules for the buses and trains
-On the other hand, a user provides her request, composed of 
-three parts.
-First, the user enters \tit{from} and \tit{to} locations 
-on the map together with day and time of the start of the trip.  
-Second, the user may upload her auxiliary metric dataset, e.g., crime rates.  
-Lastly, the user specifies her constraints in LTL and preferences as a $\PCF$.
-For example, the constraint 
-could be ``never walk through a bad neighborhood."  Given these 
-inputs, our planner computes an optimal path satisfying 
-all the constraints and optimizing the preferences.
-Note that the request from the user is encapsulated into
-a JSON object.
-
-For instance, the JSON object for the constraint $\psi$ in \eqtref{ex}
-is shown in \figref{exJSON}.
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/exampleJSON.png}
-  \caption{JSON object for constraint $\psi$ in \eqtref{ex}\label{fig:exJSON}}
-\end{figure}
-
-\subsection{Results}
-We present and analyze resulting routes for three agents:
-Alice, Bob, and Cal.
-This is assuming no auxiliary metric datasets uploaded
-so that the agents focus on time and fare.
-Fixing their where and when information, we show how
-agents' different constraints and preferences affect
-their optimal routes computed by our planner.
-Their where and when are set so that they all plan to
-travel from San Jose International Airport (SJC) in
-San Jose, to Pier 39 in San Francisco.
-Note that the natural constraint $\psi$ in \eqtref{ex} 
-is implicitly imposed on all cases.
-
-Agent Alice has only one constraint that she does not 
-have a car; thus, driving to her is never available during this trip.
-Per her preferences, Alice provides her thoughts as earlier;
-that is, public transportation is preferred to taxi, which is
-better than biking, preferred to walking.
-What's more, she decides that she would sacrifice thirty dollars
-to save one hour during the trip.
-
-The resulting route for Alice is included in \figref{alice}.
-The solution takes 2 hours 7 minutes and 18 dollars 94 cents.
-It is optimal in the sense the combined value of time
-and fare is minimal among all paths from SJC to Pier 39.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/result_Alice.pdf}
-  \caption{Resulting route for Alice\label{fig:alice}}
-\end{figure}
-
-Like Alice, Bob is constrained that he will not drive a car in his travel,
-and his dollar per hour is thirty.
-Moreover, Bob makes his mind to have some workout with his bike and
-dictates that he will bike for between one and two hours.
-He expresses his preferences: biking and public transit are the most
-preferred, next is taxi, and the least preferred is walking.
-Similarly, he has done so by answering the aforementioned elicitation questions.
-
-The result for Bob is depicted in \figref{bob}.
-It spans 2 hours 57 minutes in time with the fare of 29 dollars 17 cents.
-It is so, seemingly worse than what Alice achieved, only because Bob has
-the constraint that he will for sure bike for one to two hours, and
-the preferences that put biking the most satisfying mode.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/result_Bob.pdf}
-  \caption{Resulting route for Bob\label{fig:bob}}
-\end{figure}
-
-Agent Cal will also not use a car in the trip.
-He affirms that time is greatly valuable and one hour is measured
-to 500 dollars, although his budget is restricted to 50.
-Cal's preferences are that the most preferred are public transit
-and taxi, and the next preferred are walking and biking that
-are equivalent.
-
-Refer to \figref{cal} for the optimal path for Cal.
-This route, stretching 1 hour 48 minutes, is the most time-saving one 
-compared with the previous two, at the price of 49 dollars 91 cents.
-This is due to the constraint that Cal can spend up to 50 dollars,
-as well as his preferences and vale of time being high.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/result_Cal.pdf}
-  \caption{Resulting route for Cal\label{fig:cal}}
-\end{figure}
-
-\subsubsection{Auxiliary Metric}
-When the user of the planner is interested in metrics other than
-the ones offered already (i.e., time and fare), she might discover
-new metrics (e.g., crime rates and pollution statistics), upload
-them into the planner, and retrieve optimal plans taking these
-metrics into account.  One scenario of this approach is the following.
-
-Our agent needs to commute back home at night.
-
-\begin{figure}[!ht]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/crime.pdf}
-  \caption{Optimal route considering crime rates\label{fig:crime}}
-\end{figure}
+\section{Results}
+\input{results}
 
 \section{Conclusion and Future Work}
-In multi-modal trip planning, it is vital for the planner to enable
-users to integrate new metrics of interest into the existing map,
-and to express constraints and preferences over which optimal
-routes are computed.
-In our work, we designed and built a graph-search based, extensible
-and personalizable multi-modal trip planner that utilizes linear
-temporal logic and preferential cost
-function to model user constraints and preferences.
-
-As a note on the future work, we plan to explore techniques in planning
-for computing multiple routes that are diverse with bounded difference of costs.
-Continuing this direction, we intend to study the problem of learning the
-$\PCF$ coefficients using the observations of the decisions the user made among
-the computed paths. (It is reasonable to assume that routes picked by the user
-are preferred to others, and these observations could contribute to the learning
-problem.)
-Also interesting is to introduce traffic information into the planner
-to support real-time multi-agent concurrent trip planning.
+\input{summary}
 

### 2016-03-03 personalizability.tex
Broke paper.tex in pieces.

Personalizability consists of two aspects: constraints and preferences.
From the viewpoint of the planner,
constraints, also referred to as hard constraints, are statements that the planner
has to satisfy during the planning process; whereas preferences, also called
soft constraints, are specifications that the planner will need to optimize.
We formulated constraints using linear temporal logic (LTL) and preferences as
a preferential cost function (PCF), and implemented our planner leveraging the
widely-used graph search algorithm the A*.

\subsection{Constraints}
%\begin{itemize}
%	\setlength\itemsep{1pt}
%	\item Describe syntax and semantics of LTL in the setting of trip planning.
%	\begin{itemize}
%		\setlength\itemsep{0pt}
%		\item Specifically, we need to point out the ``after" temporal connective,
%					and we need to describe the semantics: what is a plan, why we can
%					eliminate actions and put them as state attributes instead, etc.
%	\end{itemize}
%\end{itemize}
As constraints in the setting of trip planning are often declarative and
temporal, our choice of LTL is straightforward.
We now give a brief review of linear temporal logic (LTL).
Let $f$ be a propositional formula over a finite set $L$ of Boolean variables.  
LTL formulas are defined recursively as follows.
\begin{equation}
	\varphi = f | \varphi_1 \land \varphi_2 | \varphi_1 \lor \varphi_2 | \neg \varphi | 
		\bigcirc \varphi |	\Box \varphi | \Diamond \varphi | \varphi_1 \cA \varphi_2
\end{equation}
Note that we have $\varphi_1 \cA \varphi_2$, and it means that
``$\varphi_2$ holds right after $\varphi_1$ holds."

A natural constraint in trip planning could be ``In this trip I will not drive a car 
after biking or taking the public transit."
In LTL, such constraint can be translated into an LTL formula
\begin{equation}
\label{eqt:ex}
	\psi = ((M=b) \lor (M=p)) \,\cA\, (\Box (\neg (M=c))).
\end{equation}

As the actions in trip planning is limited to taking different transportation modes,
in our definition of the semantics of LTL
these actions are subsumed into the interpretations of $L$, or \tit{states}.
The semantics of LTL is defined with regard to trajectories of states. 
Let $\sigma$ be a trajectory of states $S_0,a_1,S_1,\ldots,a_n,S_n$, and
$\sigma[i]$ a suffix $S_i, a_{i+1}, S_{i+1}, \ldots,a_n,S_n$.  We have
\begin{align*}
	\sigma \models f \;\; &\IFF \;\; S_0 \models f,\\
	\sigma \models \varphi_1 \land \varphi_2 \;\; &\IFF \;\; \sigma \models \varphi_1 \; and \; \sigma \models \varphi_2,\\
	\sigma \models \varphi_1 \lor \varphi_2 \;\; &\IFF \;\; \sigma \models \varphi_1 \; or \; \sigma \models \varphi_2,\\
	\sigma \models \neg \varphi \;\; &\IFF \;\; \sigma \not \models \varphi,\\
	\sigma \models \bigcirc \varphi \;\; &\IFF \;\; \sigma[1] \models \varphi,\\
	\sigma \models \Box \varphi \;\; &\IFF \;\; \forall 0 \leq i \leq n (\sigma[i] \models \varphi),\\
	\sigma \models \Diamond \varphi \;\; &\IFF \;\; \exists 0 \leq i \leq n (\sigma[i] \models \varphi),\\
	\sigma \models \varphi_1 \cA \varphi_2 \;\; &\IFF \;\; \forall 0 \leq i < n (\IF \; \sigma[i] \models \varphi_1, \sigma[i+1] \models \varphi_2).
\end{align*}

For example, we are given an LTL constraint $\psi$ and three trajectories ($\sigma_1$, $\sigma_2$ and $\sigma_3$)
as shown in \figref{trjs}.
Clearly, we have $\sigma_1 \models \psi$ because, after public transit in $S_2$ and $S_3$, 
traveling by car has never been taken place.
Moreover, we have $\sigma_2 \not \models \psi$ because we have $M=c$ hold in $S_6$ and $S_7$
after having $M=p$ hold in $S_2$ and $S_5$.
Finally, we know $\sigma_3 \models \psi$, as the mode is always neither biking nor public transit.

\begin{figure}[!ht]
  \centering
    \includegraphics[width=0.4\textwidth]{figs/trajectories.pdf}
  \caption{State transition diagram\label{fig:trjs}}
\end{figure}

\subsection{Preferences}
%\begin{itemize}
%	\setlength\itemsep{1pt}
%	\item Describe our preferential cost function (PCF).
%	\begin{itemize}
%		\setlength\itemsep{0pt}
%		\item Describe why and how we break down time and money.
%		\item Describe why and how we design coefficients to these time and money pieces.
%		\item Describe why and how we design the two ratios: dollars per hour and dollars per aux.
%	\end{itemize}
%\end{itemize}

A state is described as a set of \tit{state variables}.
The state variables of a state $S$ include the transportation mode $M$ that led to $S$,
time $T$ spent so far per mode (e.g., $T_\bike$ for biking and $T_\public$ for
public transit), fare $D$ spent so far per mode (e.g., $D_\gas$ for driving and
$D_\taxi$ for taking a cab), and variables related to the auxiliary data once uploaded.
These extra data related variables are metrics such as the sum ($A_\SUM$),
the maximum ($A_\MAX)$, the minimum ($A_\MIN$), and the average ($A_\AVG$) data along the path.

We focused on weighted functions over state variables and
designed the cost function, called preferential cost function (PCF), that guides the
graph-based search engine in our trip planner as follows.
\begin{equation}
	\begin{aligned}
		\PCF(S) =& \beta_1 * (\alpha_\walk \cdot T_\walk + \alpha_\bike \cdot T_\bike + \ldots) \\
								&+ (D_\gas + D_\public + \ldots) \\
								&+ \beta_2 * (A_\SUM + \ldots),
	\end{aligned}
	\label{eqt:pcf}
\end{equation}
where $\alpha_i$ are real numbers representing the relations among different time pieces,
and $\beta_1$ ($\beta_2$) is the ratio that essentially describes how much in dollars a user would pay to
save an hour (an auxiliary data, respectively).

\subsubsection{Preference Elicitation}
To gather these coefficients ($\alpha_i$'s and $\beta_i$'s) in our $\PCF$, we designed interface to
elicit these numbers from the user.
It asks the user questions and collect answers from the user to derive the coefficients.
These questions are as follows.
\begin{enumerate}
	\item One hr of walking $=$ \underline{\hspace{1cm}} hrs of driving?
	\item One hr of biking $=$ \underline{\hspace{1cm}} hrs of driving?
	\item One hr of public transit $=$ \underline{\hspace{1cm}} hrs of driving?
	\item One hr of taxi $=$ \underline{\hspace{1cm}} hrs of driving?
	\item How much in dollars would you pay to save an hour in traveling?
	\item How much in dollars would you pay to avoid an auxiliary event (e.g., crime) in traveling?
\end{enumerate}
For instance, Alice, an agent, answers 3, 2, 0.25 and 0.5 to the first four questions in the list above.
Intuitively, the numbers indicate that she prefers public transit the most, followed by
taxi, driving, biking and walking, in order.
We show how we can now derive $\alpha_i$'s in \eqtref{pcf}.
We start with setting $\alpha_\car=1$.
Now, since one hour of walking is equivalent to 3 hours of walking,
we have $\alpha_\walk \times 1 = \alpha_\car \times 3$;
hence, we derive $\alpha_\walk = 3$.
Similarly, we have $\alpha_\bike = 2$, $\alpha_\public = 0.25$,
and $\alpha_\taxi = 0.5$.
Computing $\beta_1$ and $\beta_1$ is straightforward.
Indeed, function $\PCF$ with the input of time, fare and
auxiliary metric pieces boils down to monetary cost,
and the planner computes the best path by optimize
based on this overall monetary cost in the searching
process.

\subsection{Reasoning with Constraints and Preferences}
%\begin{itemize}
%	\setlength\itemsep{1pt}
%	\item Describe why and how we integrate constraints and preferences into A*.
%\end{itemize}

\begin{figure}[!ht]
  \centering
    \includegraphics[width=0.4\textwidth]{figs/Astar.pdf}
  \caption{Adjusted A*\label{fig:astar}}
\end{figure}

We leveraged the widely-used A* search algorithm on top of our high-performance graph search
engine (cf. \figref{astar}).  The A* algorithm incorporates the following cost function.
\begin{equation}
	f(S) = g(S) + h(S),
\end{equation}
where $g(S)$ is the cost of an optimal trip from the initial state to $S$, and
$h(S)$ is an admissible estimate of the cost of an optimal trip from $S$ to goal.

To prune the search space, we check satisfiability of the temporal constraints in LTL
at expansion of the search tree.
To guide the search engine, we set $g(S)=\PCF(S)$ and $h(S)$ the minimum estimate among
all available modes in $S$.

### 2016-03-03 results.tex
Broke paper.tex in pieces.

We present and analyze resulting routes for three agents:
Alice, Bob, and Cal.
This is assuming no auxiliary metric datasets uploaded
so that the agents focus on time and fare.
Fixing their where and when information, we show how
agents' different constraints and preferences affect
their optimal routes computed by our planner.
Their where and when are set so that they all plan to
travel from San Jose International Airport (SJC) in
San Jose, to Pier 39 in San Francisco.
Note that the natural constraint $\psi$ in \eqtref{ex} 
is implicitly imposed on all cases.

Agent Alice has only one constraint that she does not 
have a car; thus, driving to her is never available during this trip.
Per her preferences, Alice provides her thoughts as earlier;
that is, public transportation is preferred to taxi, which is
better than biking, preferred to walking.
What's more, she decides that she would sacrifice thirty dollars
to save one hour during the trip.

The resulting route for Alice is included in \figref{alice}.
The solution takes 2 hours 7 minutes and 18 dollars 94 cents.
It is optimal in the sense the combined value of time
and fare is minimal among all paths from SJC to Pier 39.

\begin{figure}[!ht]
  \centering
    \includegraphics[width=0.5\textwidth]{figs/result_Alice.pdf}
  \caption{Resulting route for Alice\label{fig:alice}}
\end{figure}

Like Alice, Bob is constrained that he will not drive a car in his travel,
and his dollar per hour is thirty.
Moreover, Bob makes his mind to have some workout with his bike and
dictates that he will bike for between one and two hours.
He expresses his preferences: biking and public transit are the most
preferred, next is taxi, and the least preferred is walking.
Similarly, he has done so by answering the aforementioned elicitation questions.

The result for Bob is depicted in \figref{bob}.
It spans 2 hours 57 minutes in time with the fare of 29 dollars 17 cents.
It is so, seemingly worse than what Alice achieved, only because Bob has
the constraint that he will for sure bike for one to two hours, and
the preferences that put biking the most satisfying mode.

\begin{figure}[!ht]
  \centering
    \includegraphics[width=0.5\textwidth]{figs/result_Bob.pdf}
  \caption{Resulting route for Bob\label{fig:bob}}
\end{figure}

Agent Cal will also not use a car in the trip.
He affirms that time is greatly valuable and one hour is measured
to 500 dollars, although his budget is restricted to 50.
Cal's preferences are that the most preferred are public transit
and taxi, and the next preferred are walking and biking that
are equivalent.

Refer to \figref{cal} for the optimal path for Cal.
This route, stretching 1 hour 48 minutes, is the most time-saving one 
compared with the previous two, at the price of 49 dollars 91 cents.
This is due to the constraint that Cal can spend up to 50 dollars,
as well as his preferences and vale of time being high.

\begin{figure}[!ht]
  \centering
    \includegraphics[width=0.5\textwidth]{figs/result_Cal.pdf}
  \caption{Resulting route for Cal\label{fig:cal}}
\end{figure}

\subsection{Auxiliary Metric}
When the user of the planner is interested in metrics other than
the ones offered already (i.e., time and fare), she might discover
new metrics (e.g., crime rates and pollution statistics), upload
them into the planner, and retrieve optimal plans taking these
metrics into account.  One scenario of this approach is the following.

Our agent needs to travel without a car across San Francisco downtown at night.
For her, safety is important.
Having found the crime statistics for the area, the agent uploads
the data as a new auxiliary metric into the map.
By specifying that she will never walk through a neighborhood
with more than fifteen crimes over the last month, and that
she would sacrifice a quarter to avoid one crime incident,
the agent tells the planner to come up with a relatively safe
route. An example is shown in \figref{crime}, where the agent
needs to start at the east of downtown and travel across the area
to arrive at the west side.  
The computed path is represented by the line colored by black,
blue and green, denoting taxi, public transit and biking, respectively.
Clearly, this path routes away from crime-heavy areas and achieves
optimality in that the combined metrics -- time, fare and crime rates,
uploaded and personalized by the user -- is minimal among all possibilities.

\begin{figure}[!ht]
  \centering
    \includegraphics[width=0.5\textwidth]{figs/crime.pdf}
  \caption{Optimal route considering crime rates\label{fig:crime}}
\end{figure}

### 2016-03-03 summary.tex
Broke paper.tex in pieces.

In multi-modal trip planning, it is vital for the planner to enable
users to integrate new metrics of interest into the existing map,
and to express constraints and preferences over which optimal
routes are computed.
In our work, we designed and built a graph-search based, extensible
and personalizable multi-modal trip planner that utilizes linear
temporal logic and preferential cost
function to model user constraints and preferences.

As a note on the future work, we plan to explore techniques in planning
for computing multiple routes that are diverse with bounded difference of costs.
Continuing this direction, we intend to study the problem of learning the
$\PCF$ coefficients using the observations of the decisions the user made among
the computed paths. (It is reasonable to assume that routes picked by the user
are preferred to others, and these observations could contribute to the learning
problem.)
Also interesting is to introduce traffic information into the planner
to support real-time multi-agent concurrent trip planning.
